### 2021-02-23 submission/submission.rmd
transfer sections to a single file

---
title: "Treatment heterogeneity in comparative effectiveness of teriparatide vs bisphosphonates: multi-database cohort study"
output:
    bookdown::pdf_document2: default
    bookdown::word_document2:
        reference_docx: reference.docx
geometry: margin=1.0in
toc: false
font-size: 11pt
header-includes:
  - \usepackage{helvet}
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage[left]{lineno}
  - \linenumbers
---

# Introduction
Osteoporosis is a common condition characterised by decreased bone density and associated with increased risk for fragility fractures, which affects almost 30% of women aged $\geq50$ years (#Wright et al. 2017). It was estimated that in 2005, over 2 million incident fractures occurred in the United States, with hip fractures accounting for more than 70% of the costs. By 2025, annual fractures and costs could rise by almost 50% (#Cole et al. 2008).

Fracture prevention is thus the key focus of anti-osteoporotic therapy, reducing the disease burden both for the affected patient but also on the population level. Several pharmacological agents are available for primary and secondary prevention. The choice of anti-osteoporotic agent largely depends on history of fragility fracture and anticipated fracture risk. Oral bisphosphonates (BP) are first line treatments for postmenopausal patients with increased fracture risk in clinical guidelines (#AACE/ACE 2020) as for their favourable cost-effectiveness and safety profile. Teriparatide, a parathyroid hormone analogue administered as a daily injection, was approved by the FDA in 2002 as the first anabolic agent for treatment of severe postmenopausal osteoporosis. While clinical trials among patients with previous vertebral fractures showed a substantial risk reduction for new vertebral fracture comparing teriparatide to placebo (#Neer et al 2001) and risedronate (#Kendler et al. 2018), its efficacy on low-incident major osteoporotic fractures, especially hip fractures, is less well established: Previous randomised trials assessing hip fracture comprised only few events, thus not providing sufficient power to show differences between treatment groups when studying comparative effectiveness. A recent meta-analysis by Diez-Perez and colleagues assessed the effect of teriparatide on hip fracture, indicating a significant 80% risk reduction compared to placebo and a non-significant 46% risk reduction when compared to active controls.

With teriparatide-containing biosimilars being launched in recent years, treatment costs dropped significantly. Subsequently, the discussion of cost-effectiveness in patients with less severe forms of osteoporosis is restarting. Therefore, evaluation of the effectiveness in hip fracture prevention is required. In particular, observational studies assessing comparative effectiveness in a real-world setting are required, as effectiveness may differ from clinical trial as for differences in patient’s persistence and adherence to their anti-osteoporotic treatment. While hip fracture is a comparatively rare outcome, it is unambiguously defined and reliably recorded in routinely collected data, allowing for a direct comparison to results from clinical trial meta-analyses. While results from non-controlled, observational studies suggest that teriparatide may reduce the risk of hip fractures (#Burge et al. 2017, #Silverman et al. 2019), this is the first observational study assessing teriparatide in hip fracture risk reduction compared to oral bisphosphonate users in postmenopausal women using multiple large real-world databases.

# Methods
## Study design
We ran our analyses on two US observational administrative claims databases and one electronic health record (EHR) database, all mapped to OMOP-CDM version. More specifically, the considered claims databases were IBM MarketScan Medicare Supplemental Database (MDCR) representing retirees in the United States with primary or Medicare supplemental coverage and Optum De-Identified Clinformatics® Data Mart Database – Date of Death (DOD), including members of private health insurance. The EHR database was Optum de-identified Electronic Health Record Dataset (PANTHER) including Humedica’s Electronic Health Record.

Our cohorts included female new users of teriparatide or oral bisphosphonates above the age of 65 (ref 7,8 LEGEND-HTN). Patients were considered to be new-users if they received no anti-osteoporosis drugs (raloxifene, bazedoxifene, denosumab, abaloparatide, romosozumab) 365 days prior to treatment initiation with teriparatide or an oral bisphosphonate. We required patients to have at least one year of observation period before first exposure and an established osteoporosis diagnosis, defined as any recorded fracture of the hip, wrist, spine or shoulder/humerus.

Our primary efficacy outcome was hip fracture. Vertebral fracture and a composite major osteoporotic fracture, defined as hip, vertebral or wrist/forearm/proximal humerus fracture, were our secondary efficacy outcomes. We excluded patients that had experienced the outcome under study any time prior to treatment initiation. Patient time-at-risk started at 1 day after treatment initiation until 730 days after treatment initiation.

# Statistical analyses
We performed two sets of analyses. First, we derived overall treatment effect estimates of teriparatide compared to oral bisphosphonates regarding the three outcomes of interest. To account for potential measured confounding we developed separate propensity score models within each database using the same predefined set of measured covariates. Covariates were included in the propensity score model using regularized logistic regression (ref Tian-Schuemie simulation). We estimated treatment effects using Cox proportional hazards models with treatment as the sole covariate fitted in the 1-4 propensity score-matched subset of the considered patient population. Results from different databases were summarized using random effects meta-analysis (ref).

For the second set of analyses, we used the standardized framework for risk-based assessment of treatment effect heterogeneity, that follows the directives of the PATH statement (ref framework). Following data extraction, the framework was applied in two steps. First, we derived individualized risk predictions for the three efficacy outcomes. We built the prediction models using LASSO logistic regression on the propensity score matched (1 to 4) subpopulation of the pooled treatment arms, aiming to remove any effect of measured confounding that may cause differential fit of our models across treatment arms. We considered the same large set of candidate covariates for the development of our prediction models. Time at risk was set at 730 days after study inclusion.

For each outcome we used the derived prediction models to divide the population under study in 4 equally-sized subgroups, based on the quarters of the predicted risk distribution. Within each of these risk-based subgroups we developed a new propensity score model. Our analyses were performed on the propensity score matched (1 to 4) subset of the risk subgroup subset. We derived relative effect estimates using Cox proportional hazards models only with treatment as a predictor. Absolute effect estimates were calculated based on the difference of the Kaplan-Meier estimates.

# Results
A total of 758,879 patients were initially considered for our analyses. Of those, 18,809 received teriparatide and 740,070 received an oral bisphosphonate. Table 1 presents the patients considered in the final analyses across the 3 databases, after excluding patients experiencing the outcomes under study any time prior to cohort inclusion.

The random effects meta-analysis of teriparatide compared to oral bisphosphonates across the 3 considered databases gave an overall hazard ratio of 0.93 (0.78 to 1.10; 95% CI) for hip fracture, 1.09 (1.01 to 1.17; 95%CI) for major osteoporotic fracture and 1.00 (0.90 to 1.12; 95%CI) for vertebral fracture (Figures 1-3).

For the assessment of treatment effect heterogeneity none of the internally developed prediction models achieved high discrimination, with the models predicting the composite major osteoporotic fracture suffering the most (Table 2). However, models on hip fracture—the primary efficacy outcome—displayed adequate performance. More information on the performance of the prediction models can be found in the supplement.

When analyzing the absolute risk difference of teriparatide compared to oral bisphosphonates within strata of predicted hip fracture risk we did not observe any clear pattern except for a non-significant absolute benefit for hip fracture within the highest risk patients (Figure 4), namely 0.55% (-0.89 to 1.98; 95%CI) in MDCR, 0.97% (-0.27 to 2.22; 95%CI) in OPTUM_DOD and 0.45% (-0.56 to 1.46; 95%CI) in PANTHER. Absolute and relative treatment effect estimates within strata of predicted risk of the secondary outcomes are available in the supplement.

### 2021-02-23 submission/manuscript.rmd
fix naming to manuscript

--- 2021-02-23 submission/submission.rmd
@@
   - \linenumbers
+editor_options: 
+  chunk_output_type: console
 ---
+
+```{r, echo=FALSE, message=FALSE}
+library(tidyverse)
+```
 
@@
 Our analyses were performed on the propensity score matched (1 to 4) subset of the risk subgroup subset.
+We used a set of XXX negative control outcomes, to calibrate our results for unmeasured confounding (##REFEFRENCE).
 We derived relative effect estimates using Cox proportional hazards models only with treatment as a predictor.
@@
 # Results
-A total of 758,879 patients were initially considered for our analyses.
-Of those, 18,809 received teriparatide and 740,070 received an oral bisphosphonate.
+
+```{r, echo=FALSE}
+incidenceOverall <- readRDS(here::here("data/raw/incidenceOverall.rds")) %>%
+  dplyr::filter(analysisType == "matchOnPs_1_to_4") %>%
+  dplyr::group_by(database) %>%
+  dplyr::summarise(
+    maxTreatment = max(treatmentPersons),
+    maxComparator = max(comparatorPersons),
+    .groups = "drop"
+  )
+```
+
+<!-- Need to fix the numbers. See icpe abstract -->
+A total of `r sum(incidenceOverall$maxTreatment) + sum(incidenceOverall$maxComparator)` patients were initially considered for our analyses.
+Of those, `r sum(incidenceOverall$maxTreatment)` received teriparatide and `r sum(incidenceOverall$maxComparator)` received an oral bisphosphonate.
 Table 1 presents the patients considered in the final analyses across the 3 databases, after excluding patients experiencing the outcomes under study any time prior to cohort inclusion.
+
+```{r, echo=FALSE}
+readRDS(here::here("data/raw/mappedOverallResultsNegativeControls.rds")) %>%
+  dplyr::filter(analysisType == "matchOnPs_1_to_4") %>%
+  head()
+```
 

### 2021-02-23 code/MetaAnalysis.R
start calibration R script

@@ -1,0 +1,182 @@
+library(dplyr)
+
+# filePath <- "/home/arekkas/Documents/Projects/osteoporosis/meta"
+# files <- list.files(
+#   path = filePath,
+#   full.names = TRUE
+# )
+# 
+# res <- data.frame()
+# for (file in seq_along(files)) {
+#   res1 <- readRDS(files[file])
+#   res <- rbind(res, res1)
+# }
+# 
+
+overallNegativeControls <- readRDS(
+  "/home/arekkas/Documents/Projects/osteoporosis/multipleRseeAnalyses/Data/mappedOverallResultsNegativeControls.rds"
+)
+overallMappedOverallRelativeResults <- readRDS(
+  "/home/arekkas/Documents/Projects/osteoporosis/multipleRseeAnalyses/Data/mappedOverallResults.rds"
+)
+databases <- unique(overallNegativeControls$database)
+
+res <- data.frame(
+  logRr = numeric(),
+  logLb95Rr = numeric(),
+  logUb95Rr = numeric(),
+  seLogRr = numeric(),
+  database = character()
+)
+
+for (i in seq_along(databases)) {
+  ncs <- overallNegativeControls %>% 
+    filter(analysisType == "matchOnPs_1_to_4", database == databases[i]) %>%
+    mutate(logRr = log(estimate))
+  
+  mod <- EmpiricalCalibration::fitSystematicErrorModel(
+    logRr =  ncs$logRr,
+    seLogRr = ncs$seLogRr,
+    trueLogRr = rep(0, nrow(ncs))
+  )
+  est <- overallMappedOverallRelativeResults %>%
+    filter(
+      analysisType == "matchOnPs_1_to_4",
+      outcomeId == 101,
+      database == database[i]
+      ) %>%
+    mutate(logRr = log(estimate))
+  
+  res <- res %>%
+    dplyr::bind_rows(
+    EmpiricalCalibration::calibrateConfidenceInterval(
+      logRr = est$logRr,
+      seLogRr = est$seLogRr,
+      model = mod
+    ) %>%
+      dplyr::mutate(
+        database = databases[i]
+      )
+  )
+  
+}
+
+# analyze <- res %>%
+#   # dplyr::filter(
+#     # outcomeId == 101,
+#     # analysisType == "matchOnPs_1_to_4",
+#   # ) %>%
+#   dplyr::select(
[…]

### 2021-02-23 code/calibrate.R

@@ -1,0 +1,87 @@
+library(tidyverse)
+
+overallNegativeControls <- readRDS(
+  "/home/arekkas/Documents/Projects/osteoporosis/data/raw/mappedOverallResultsNegativeControls.rds"
+) %>%
+    dplyr::filter(
+        analysisType == "matchOnPs_1_to_4"
+    ) %>%
+    dplyr::mutate(logRr = log(estimate))
+
+overallMappedOverallRelativeResults <- readRDS(
+  "/home/arekkas/Documents/Projects/osteoporosis/data/raw/mappedOverallResults.rds"
+) %>%
+    dplyr::filter(
+        analysisType == "matchOnPs_1_to_4"
+    ) %>%
+    mutate(logRr = log(estimate))
+
+mod <- overallNegativeControls %>%
+    split(.$database) %>%
+    purrr::map(
+        ~EmpiricalCalibration::fitSystematicErrorModel(
+            logRr = .x$logRr,
+            seLogRr = .x$seLogRr,
+            trueLogRr = rep(0, nrow(.x))
+        )
+    )
+
+overallMappedOverallRelativeResults %>%
+    split(list(.$database, .$outcomeId), sep = ":") %>%
+    purrr::map_dfr(
+        ~EmpiricalCalibration::calibrateConfidenceInterval(
+            logRr = .x$logRr,
+            seLogRr = .x$seLogRr,
+            model = mod[[.x$database]]
+        ),
+        .id = "id"
+    ) %>%
+    mutate(
+        hr = exp(logRr),
+        lower = exp(logLb95Rr),
+        upper = exp(logUb95Rr)
+    ) %>%
+    tidyr::separate(id, c("database", "outcome"), ":") %>%
+    saveRDS("data/processed/calibrateOverallResults.rds")
+
+
+res <- data.frame(
+  logRr = numeric(),
+  logLb95Rr = numeric(),
+  logUb95Rr = numeric(),
+  seLogRr = numeric(),
+  database = character()
+)
+
+for (i in seq_along(databases)) {
+  ncs <- overallNegativeControls %>% 
+    filter(analysisType == "matchOnPs_1_to_4", database == databases[i]) %>%
+    mutate(logRr = log(estimate))
+  
+  mod <- EmpiricalCalibration::fitSystematicErrorModel(
+    logRr =  ncs$logRr,
+    seLogRr = ncs$seLogRr,
+    trueLogRr = rep(0, nrow(ncs))
+  )
[…]

### 2021-02-24 submission/manuscript.rmd
inline code for overall meta-analyses

@@
 library(tidyverse)
+library(knitr)
+options(scipen = 1, digits = 2)
+knit_hooks$set(
+  inline = function(x) {
+    prettyNum(
+      x, 
+      big.mark = ",",
+      decimal.mark = ".",
+      preserve.width = "individual"
+    )
+  }
+)
 ```
@@
 To account for potential measured confounding we developed separate propensity score models within each database using the same predefined set of measured covariates.
-Covariates were included in the propensity score model using regularized logistic regression (ref Tian-Schuemie simulation).
+Covariates were included in the propensity score model using regularized logistic regression (##REFERENCE Tian-Schuemie simulation).
 We estimated treatment effects using Cox proportional hazards models with treatment as the sole covariate fitted in the 1-4 propensity score-matched subset of the considered patient population.
-Results from different databases were summarized using random effects meta-analysis (ref).
+Results from different databases were summarized using random effects meta-analysis (##REFERECNE).
 
-For the second set of analyses, we used the standardized framework for risk-based assessment of treatment effect heterogeneity, that follows the directives of the PATH statement (ref framework).
+For the second set of analyses, we used the standardized framework for risk-based assessment of treatment effect heterogeneity, that follows the directives of the PATH statement (##REFERENCE framework).
 Following data extraction, the framework was applied in two steps.
@@
 
-For each outcome we used the derived prediction models to divide the population under study in 4 equally-sized subgroups, based on the quarters of the predicted risk distribution.
+For each outcome we used the derived prediction models to divide th e population under study in 4 equally-sized subgroups, based on the quarters of the predicted risk distribution.
 Within each of these risk-based subgroups we developed a new propensity score model.
@@
 ```{r, echo=FALSE}
-readRDS(here::here("data/raw/mappedOverallResultsNegativeControls.rds")) %>%
-  dplyr::filter(analysisType == "matchOnPs_1_to_4") %>%
-  head()
+metaCalibrateOverall <- readRDS(
+  here::here("data/processed/metaCalibrateOverall.rds")
+)
 ```
 
-The random effects meta-analysis of teriparatide compared to oral bisphosphonates across the 3 considered databases gave an overall hazard ratio of 0.93 (0.78 to 1.10; 95% CI) for hip fracture, 1.09 (1.01 to 1.17; 95%CI) for major osteoporotic fracture and 1.00 (0.90 to 1.12; 95%CI) for vertebral fracture (Figures 1-3).
+The random effects meta-analysis of teriparatide compared to oral bisphosphonates across the 3 considered databases gave an overall hazard ratio of `r metaCalibrateOverall %>% filter(outcome == 101) %>% pull(hr)` (`r metaCalibrateOverall %>% filter(outcome == 101) %>% pull(lower)` to `r metaCalibrateOverall %>% filter(outcome == 101) %>% pull(upper)`; 95% CI) for hip fracture, `r metaCalibrateOverall %>% filter(outcome == 102) %>% pull(hr)` (`r metaCalibrateOverall %>% filter(outcome == 102) %>% pull(lower)` to `r metaCalibrateOverall %>% filter(outcome == 102) %>% pull(upper)`; 95%CI) for major osteoporotic fracture and `r metaCalibrateOverall %>% filter(outcome == 103) %>% pull(hr)` (`r metaCalibrateOverall %>% filter(outcome == 103) %>% pull(lower)` to `r metaCalibrateOverall %>% filter(outcome == 103) %>% pull(upper)`; 95%CI) for vertebral fracture (Figures 1-3).
 
-For the assessment of treatment effect heterogeneity none of the internally developed prediction models achieved high discrimination, with the models predicting the composite major osteoporotic fracture suffering the most (Table 2).
-However, models on hip fracture—the primary efficacy outcome—displayed adequate performance.
+For the assessment of treatment effect heterogeneity none of the internally developed prediction models achieved high discrimination, with the models predicting the composite major osteoporotic fracture suffering the most (##Table 2).
+However, models on hip fracture--the primary efficacy outcome--displayed adequate performance.
 More information on the performance of the prediction models can be found in the supplement.

### 2021-02-24 code/MetaAnalysis.R

@@ -1,182 +1,0 @@
-library(dplyr)
-
-# filePath <- "/home/arekkas/Documents/Projects/osteoporosis/meta"
-# files <- list.files(
-#   path = filePath,
-#   full.names = TRUE
-# )
-# 
-# res <- data.frame()
-# for (file in seq_along(files)) {
-#   res1 <- readRDS(files[file])
-#   res <- rbind(res, res1)
-# }
-# 
-
-overallNegativeControls <- readRDS(
-  "/home/arekkas/Documents/Projects/osteoporosis/multipleRseeAnalyses/Data/mappedOverallResultsNegativeControls.rds"
-)
-overallMappedOverallRelativeResults <- readRDS(
-  "/home/arekkas/Documents/Projects/osteoporosis/multipleRseeAnalyses/Data/mappedOverallResults.rds"
-)
-databases <- unique(overallNegativeControls$database)
-
-res <- data.frame(
-  logRr = numeric(),
-  logLb95Rr = numeric(),
-  logUb95Rr = numeric(),
-  seLogRr = numeric(),
-  database = character()
-)
-
-for (i in seq_along(databases)) {
-  ncs <- overallNegativeControls %>% 
-    filter(analysisType == "matchOnPs_1_to_4", database == databases[i]) %>%
-    mutate(logRr = log(estimate))
-  
-  mod <- EmpiricalCalibration::fitSystematicErrorModel(
-    logRr =  ncs$logRr,
-    seLogRr = ncs$seLogRr,
-    trueLogRr = rep(0, nrow(ncs))
-  )
-  est <- overallMappedOverallRelativeResults %>%
-    filter(
-      analysisType == "matchOnPs_1_to_4",
-      outcomeId == 101,
-      database == database[i]
-      ) %>%
-    mutate(logRr = log(estimate))
-  
-  res <- res %>%
-    dplyr::bind_rows(
-    EmpiricalCalibration::calibrateConfidenceInterval(
-      logRr = est$logRr,
-      seLogRr = est$seLogRr,
-      model = mod
-    ) %>%
-      dplyr::mutate(
-        database = databases[i]
-      )
-  )
-  
-}
-
-# analyze <- res %>%
-#   # dplyr::filter(
-#     # outcomeId == 101,
-#     # analysisType == "matchOnPs_1_to_4",
-#   # ) %>%
-#   dplyr::select(
[…]

### 2021-02-24 code/calibrate.R

@@ -1,87 +1,0 @@
-library(tidyverse)
-
-overallNegativeControls <- readRDS(
-  "/home/arekkas/Documents/Projects/osteoporosis/data/raw/mappedOverallResultsNegativeControls.rds"
-) %>%
-    dplyr::filter(
-        analysisType == "matchOnPs_1_to_4"
-    ) %>%
-    dplyr::mutate(logRr = log(estimate))
-
-overallMappedOverallRelativeResults <- readRDS(
-  "/home/arekkas/Documents/Projects/osteoporosis/data/raw/mappedOverallResults.rds"
-) %>%
-    dplyr::filter(
-        analysisType == "matchOnPs_1_to_4"
-    ) %>%
-    mutate(logRr = log(estimate))
-
-mod <- overallNegativeControls %>%
-    split(.$database) %>%
-    purrr::map(
-        ~EmpiricalCalibration::fitSystematicErrorModel(
-            logRr = .x$logRr,
-            seLogRr = .x$seLogRr,
-            trueLogRr = rep(0, nrow(.x))
-        )
-    )
-
-overallMappedOverallRelativeResults %>%
-    split(list(.$database, .$outcomeId), sep = ":") %>%
-    purrr::map_dfr(
-        ~EmpiricalCalibration::calibrateConfidenceInterval(
-            logRr = .x$logRr,
-            seLogRr = .x$seLogRr,
-            model = mod[[.x$database]]
-        ),
-        .id = "id"
-    ) %>%
-    mutate(
-        hr = exp(logRr),
-        lower = exp(logLb95Rr),
-        upper = exp(logUb95Rr)
-    ) %>%
-    tidyr::separate(id, c("database", "outcome"), ":") %>%
-    saveRDS("data/processed/calibrateOverallResults.rds")
-
-
-res <- data.frame(
-  logRr = numeric(),
-  logLb95Rr = numeric(),
-  logUb95Rr = numeric(),
-  seLogRr = numeric(),
-  database = character()
-)
-
-for (i in seq_along(databases)) {
-  ncs <- overallNegativeControls %>% 
-    filter(analysisType == "matchOnPs_1_to_4", database == databases[i]) %>%
-    mutate(logRr = log(estimate))
-  
-  mod <- EmpiricalCalibration::fitSystematicErrorModel(
-    logRr =  ncs$logRr,
-    seLogRr = ncs$seLogRr,
-    trueLogRr = rep(0, nrow(ncs))
-  )
[…]

### 2021-02-24 code/calibrateOverall.R

@@ -1,0 +1,58 @@
+#!/usr/bin/env Rscript
+
+# Description:
+#   Calculates the negative control-calibrated hazard ratios per database:outcome
+#   combination.
+# Depends: 
+#   data/raw/mappedOverallResultsNegativeControls.rds
+#   data/raw/mappedOverallResults.rds
+# Output:
+#   data/processed/calibrateOverallResults.rds
+
+
+library(tidyverse)
+
+overallNegativeControls <- readRDS(
+  "/home/arekkas/Documents/Projects/osteoporosis/data/raw/mappedOverallResultsNegativeControls.rds"
+) %>%
+  dplyr::filter(
+    analysisType == "matchOnPs_1_to_4"
+  ) %>%
+  dplyr::mutate(logRr = log(estimate))
+
+overallMappedOverallRelativeResults <- readRDS(
+  "/home/arekkas/Documents/Projects/osteoporosis/data/raw/mappedOverallResults.rds"
+) %>%
+  dplyr::filter(
+    analysisType == "matchOnPs_1_to_4"
+  ) %>%
+  mutate(logRr = log(estimate))
+
+mod <- overallNegativeControls %>%
+  split(.$database) %>%
+  purrr::map(
+    ~EmpiricalCalibration::fitSystematicErrorModel(
+      logRr = .x$logRr,
+      seLogRr = .x$seLogRr,
+      trueLogRr = rep(0, nrow(.x))
+    )
+  )
+
+
+overallMappedOverallRelativeResults %>%
+  split(list(.$database, .$outcomeId), sep = ":") %>%
+  purrr::map_dfr(
+    ~EmpiricalCalibration::calibrateConfidenceInterval(
+      logRr = .x$logRr,
+      seLogRr = .x$seLogRr,
+      model = mod[[.x$database]]
+    ),
+    .id = "id"
+  ) %>%
+  mutate(
+    hr = exp(logRr),
+    lower = exp(logLb95Rr),
+    upper = exp(logUb95Rr)
+  ) %>%
+  tidyr::separate(id, c("database", "outcome"), ":") %>%
+  saveRDS("data/processed/calibrateOverallResults.rds")

### 2021-02-24 code/metaCalibrateOverall.R

@@ -1,0 +1,43 @@
+#!/usr/bin/env Rscript
+
+# Description:
+#     Generates meta-analytic estimates of negative control-calibrated estimates
+#     for all outcomes
+# Depends:
+#     data/processed/calibrateOverallResults.rds
+# Output:
+#     data/processed/calibrateMetaOverall.rds
+
+
+library(tidyverse)
+
+calibrateOverallResults <- readRDS(
+    here::here("data/processed/calibrateOverallResults.rds")
+)
+
+metaAnalysis <- function(data) {
+    metaRes <- meta::metagen(
+        TE = data$logRr,
+        seTE = data$seLogRr,
+        sm = "HR",
+        method.tau = "PM",
+        studlab = data$database
+    )
+
+   res <- tibble(
+       hr = exp(metaRes$TE.random),
+       lower = exp(metaRes$lower.random),
+       upper = exp(metaRes$upper.random)
+   ) 
+   
+   return(res)
+    
+}
+
+calibrateOverallResults %>%
+    group_by(outcome) %>%
+    nest() %>%
+    mutate(var = map(data, ~metaAnalysis(.x))) %>%
+    unnest(cols = var) %>%
+    select(-data) %>%
+    saveRDS("data/processed/metaCalibrateOverall.rds")

### 2021-02-24 submission/manuscript.rmd
incorporate numeric results

@@
 
+\newpage
 # Introduction
@@
 
+\newpage
 # Methods
@@
 
-# Statistical analyses
+##     Statistical analyses
 We performed two sets of analyses.
@@
 We considered the same large set of candidate covariates for the development of our prediction models.
-Time at risk was set at 730 days after study inclusion.
+Time at risk was set at 1 to 730 days after study inclusion.
 
@@
 
+\newpage
 # Results
@@
 A total of `r sum(incidenceOverall$maxTreatment) + sum(incidenceOverall$maxComparator)` patients were initially considered for our analyses.
-Of those, `r sum(incidenceOverall$maxTreatment)` received teriparatide and `r sum(incidenceOverall$maxComparator)` received an oral bisphosphonate.
-Table 1 presents the patients considered in the final analyses across the 3 databases, after excluding patients experiencing the outcomes under study any time prior to cohort inclusion.
+Of those, `r sum(incidenceOverall$maxTreatment)` received teriparatide and `r sum(incidenceOverall$maxComparator)` received an oral bisphosphonate (##TABLE1).
 
@@
 
-The random effects meta-analysis of teriparatide compared to oral bisphosphonates across the 3 considered databases gave an overall hazard ratio of `r metaCalibrateOverall %>% filter(outcome == 101) %>% pull(hr)` (`r metaCalibrateOverall %>% filter(outcome == 101) %>% pull(lower)` to `r metaCalibrateOverall %>% filter(outcome == 101) %>% pull(upper)`; 95% CI) for hip fracture, `r metaCalibrateOverall %>% filter(outcome == 102) %>% pull(hr)` (`r metaCalibrateOverall %>% filter(outcome == 102) %>% pull(lower)` to `r metaCalibrateOverall %>% filter(outcome == 102) %>% pull(upper)`; 95%CI) for major osteoporotic fracture and `r metaCalibrateOverall %>% filter(outcome == 103) %>% pull(hr)` (`r metaCalibrateOverall %>% filter(outcome == 103) %>% pull(lower)` to `r metaCalibrateOverall %>% filter(outcome == 103) %>% pull(upper)`; 95%CI) for vertebral fracture (Figures 1-3).
+The random effects meta-analysis of teriparatide compared to oral bisphosphonates across the 3 considered databases gave a calibrated overall hazard ratio of `r metaCalibrateOverall %>% filter(outcome == 101) %>% pull(hr)` (`r metaCalibrateOverall %>% filter(outcome == 101) %>% pull(lower)` to `r metaCalibrateOverall %>% filter(outcome == 101) %>% pull(upper)`; 95% CI) for hip fracture, `r metaCalibrateOverall %>% filter(outcome == 102) %>% pull(hr)` (`r metaCalibrateOverall %>% filter(outcome == 102) %>% pull(lower)` to `r metaCalibrateOverall %>% filter(outcome == 102) %>% pull(upper)`; 95%CI) for major osteoporotic fracture and `r metaCalibrateOverall %>% filter(outcome == 103) %>% pull(hr)` (`r metaCalibrateOverall %>% filter(outcome == 103) %>% pull(lower)` to `r metaCalibrateOverall %>% filter(outcome == 103) %>% pull(upper)`; 95%CI) for vertebral fracture (##Figures 1-3).
 
-For the assessment of treatment effect heterogeneity none of the internally developed prediction models achieved high discrimination, with the models predicting the composite major osteoporotic fracture suffering the most (##Table 2).
+```{r, echo=FALSE} 
+metaCalibrateRiskStratified <- readRDS(
+  here::here("data/processed/metaCalibrateRiskStratified.rds")
+)
+
+calibrateRiskStratified <- readRDS(
+  here::here("data/processed/calibrateRiskStratified.rds")
+) %>%
+  mutate(
+    hr = exp(logRr),
+    lower = exp(logLb95Rr),
+    upper = exp(logUb95Rr)
+  )
+
+hipFractureAbsolute <- readRDS(
+  here::here("data/processed/hipFractureAbsolute.rds")
+) %>%
+  filter(riskStratum == "Q4") %>%
+  mutate(
+    estimate = 100 * estimate,
+    lower    = 100 * lower,
+    upper    = 100 * upper
+  )
+
+```
+
+When assessing heterogeneity of treatment effect, we estimated calibrated hazard ratios of `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "mdcr") %>% pull(hr)` (`r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "mdcr") %>% pull(lower)` to `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "mdcr") %>% pull(upper)`; 95% CI),
+`r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "optum_dod") %>% pull(hr)` (`r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "optum_dod") %>% pull(lower)` to `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "optum_dod") %>% pull(upper)`; 95% CI) and `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "panther") %>% pull(hr)` (`r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "panther") %>% pull(hr)` to `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "panther") %>% pull(hr)`; 95% CI) within the highest risk quarter of MDCR, OPTUM_DOD and PANTHER respectively.
+This translated in absolute risk reduction of `r hipFractureAbsolute %>% filter(database == "mdcr") %>% pull(estimate)`% (`r hipFractureAbsolute %>% filter(database == "mdcr") %>% pull(lower)`% to `r hipFractureAbsolute %>% filter(database == "mdcr") %>% pull(upper)`%; 95% CI), `r hipFractureAbsolute %>% filter(database == "optum_dod") %>% pull(estimate)`% (`r hipFractureAbsolute %>% filter(database == "optum_dod") %>% pull(lower)`% to `r hipFractureAbsolute %>% filter(database == "optum_dod") %>% pull(upper)`%; 95% CI), `r hipFractureAbsolute %>% filter(database == "panther") %>% pull(estimate)`% (`r hipFractureAbsolute %>% filter(database == "panther") %>% pull(lower)`% to `r hipFractureAbsolute %>% filter(database == "panther") %>% pull(upper)`%; 95% CI).
+The random effects meta-analytic calibrated hazard ratio within the highest risk quarter was `r metaCalibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101) %>% pull(hr)` (`r metaCalibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101) %>% pull(lower)` to `r metaCalibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101) %>% pull(upper)`; 95% CI) across the 3 databases.
+
+<!-- How much sense does it make to look at meta-analytic hazard ratios in risk-quarters? Different models were used in each case  -->
+
+<!-- For the assessment of treatment effect heterogeneity none of the internally developed prediction models achieved high discrimination, with the models predicting the composite major osteoporotic fracture suffering the most (##Table 2).
 However, models on hip fracture--the primary efficacy outcome--displayed adequate performance.
 More information on the performance of the prediction models can be found in the supplement.
+-->
 
-When analyzing the absolute risk difference of teriparatide compared to oral bisphosphonates within strata of predicted hip fracture risk we did not observe any clear pattern except for a non-significant absolute benefit for hip fracture within the highest risk patients (Figure 4), namely 0.55% (-0.89 to 1.98; 95%CI) in MDCR, 0.97% (-0.27 to 2.22; 95%CI) in OPTUM_DOD and 0.45% (-0.56 to 1.46; 95%CI) in PANTHER.
+<!-- When analyzing the absolute risk difference of teriparatide compared to oral bisphosphonates within strata of predicted hip fracture risk we did not observe any clear pattern except for a non-significant absolute benefit for hip fracture within the highest risk patients (Figure 4), namely 0.55% (-0.89 to 1.98; 95%CI) in MDCR, 0.97% (-0.27 to 2.22; 95%CI) in OPTUM_DOD and 0.45% (-0.56 to 1.46; 95%CI) in PANTHER.
 Absolute and relative treatment effect estimates within strata of predicted risk of the secondary outcomes are available in the supplement.
+-->

### 2021-02-24 code/calibrateOverall.R

@@ -13,7 +13,7 @@
 library(tidyverse)
 
 overallNegativeControls <- readRDS(
-  "/home/arekkas/Documents/Projects/osteoporosis/data/raw/mappedOverallResultsNegativeControls.rds"
+  "data/raw/mappedOverallResultsNegativeControls.rds"
 ) %>%
   dplyr::filter(
     analysisType == "matchOnPs_1_to_4"
@@ -21,7 +21,7 @@ overallNegativeControls <- readRDS(
   dplyr::mutate(logRr = log(estimate))
 
 overallMappedOverallRelativeResults <- readRDS(
-  "/home/arekkas/Documents/Projects/osteoporosis/data/raw/mappedOverallResults.rds"
+  "data/raw/mappedOverallResults.rds"
 ) %>%
   dplyr::filter(
     analysisType == "matchOnPs_1_to_4"

### 2021-02-24 code/calibrateRiskStratified.R

@@ -1,0 +1,66 @@
+#!/usr/bin/env Rscript
+
+# Description:
+#   Calculates the negative control-calibrated hazard ratios per database:outcome
+#   combination.
+# Depends: 
+#   data/raw/negativeControls.rds
+#   data/raw/mappedOverallRelativeResults.rds
+# Output:
+#   data/processed/calibrateRiskStratified.rds
+
+
+library(tidyverse)
+
+negativeControls <- readRDS(
+    "data/raw/negativeControls.rds"
+) %>%
+    dplyr::filter(
+        analysisType == "matchOnPs_1_to_4"
+    ) %>%
+    dplyr::mutate(logRr = log(HR))
+
+relativeResults <- readRDS(
+        "data/raw/mappedOverallRelativeResults.rds"
+    ) %>%
+    dplyr::filter(
+        analysisType == "matchOnPs_1_to_4"
+    ) %>%
+    mutate(logRr = log(estimate))
+
+
+
+systErrorModels <- negativeControls %>%
+    filter(!is.na(seLogRr)) %>%
+    group_by(database, stratOutcome, riskStratum) %>%
+    nest() %>%
+    mutate(
+        mod = map(
+            data,
+            ~EmpiricalCalibration::fitSystematicErrorModel(
+                logRr     = .x$logRr,
+                seLogRr   = .x$seLogRr,
+                trueLogRr = rep(0, nrow(.x))
+            )
+        )
+    ) %>%
+    select(-data)
+
+tibble(relativeResults) %>%
+    mutate(estimate = log(estimate)) %>%
+    inner_join(systErrorModels) %>%
+    group_by(database, stratOutcome, estOutcome, riskStratum) %>%
+    nest() %>%
+    mutate(
+        pp = map(
+            data,
+            ~EmpiricalCalibration::calibrateConfidenceInterval(
+                logRr   = .x$estimate,
+                seLogRr = .x$seLogRr,
+                model   = .x$mod[[1]]
+            )
+        )
+    ) %>% 
+    unnest(pp) %>%
+    select(-data) %>%
+    saveRDS("data/processed/calibrateRiskStratified.rds")

### 2021-02-24 code/extractAbsoluteHip.R

@@ -1,0 +1,23 @@
+#!/usr/bin/env Rscript
+
+# Description:
+#     Extracts the subset of analyses that estimated the absolute effects for hip
+#     fracture within strata of predicted hip fracture risk
+# Depends:
+#     data/raw/mappedOverallAbsoluteResults.rds
+# Output:
+#     data/processed/hipFractureAbsolute.rds
+    
+    
+library(tidyverse)
+
+absoluteResults <- readRDS("data/raw/mappedOverallAbsoluteResults.rds") %>%
+    tibble()
+
+absoluteResults %>%
+    filter(
+        analysisType == "matchOnPs_1_to_4",
+        stratOutcome == 101,
+        estOutcome == 101
+    ) %>%
+    saveRDS("data/processed/hipFractureAbsolute.rds")

### 2021-02-24 code/metaCalibrateRiskStratified.R

@@ -1,0 +1,45 @@
+#!/usr/bin/env Rscript
+
+# Description:
+#     Generates meta-analytic estimates of negative control-calibrated estimates
+#     for all outcomes in all risk quarters
+# Depends:
+#     data/processed/calibrateRiskStratified.rds
+# Output:
+#     data/processed/metaCalibrateRiskStratified.rds
+
+library(tidyverse)
+
+calibrateRiskStratified <- readRDS("data/processed/calibrateRiskStratified.rds")
+
+metaAnalysisRiskStratified <- function(data) {
+    
+    metaRes <- meta::metagen(
+        TE         = data$logRr,
+        seTE       = data$seLogRr,
+        sm         = "HR",
+        method.tau = "PM",
+        studlab    = data$database
+    )
+    
+    res <- tibble(
+        hr    = exp(metaRes$TE.random),
+        lower = exp(metaRes$lower.random),
+        upper = exp(metaRes$upper.random)
+    ) 
+    
+    return(res)
+}
+
+calibrateRiskStratified %>%
+    group_by(stratOutcome, estOutcome, riskStratum) %>%
+    nest() %>%
+    mutate(
+        meta = map(
+            data,
+           ~metaAnalysisRiskStratified(.x) 
+        )
+    ) %>%
+    unnest(meta) %>%
+    select(-data) %>%
+    saveRDS("data/processed/metaCalibrateRiskStratified.rds")

### 2021-02-27 code/plotMetaOverall.R
make calibrated overall meta plot

@@ -1,0 +1,105 @@
+#!/usr/bin/env Rscript
+
+# Description:
+#   Generates the plot regarding the meta-analysis of the calibrated hazard
+#   ratios for the three outcomes
+# Depends:
+#   data/processed/calibrateOverallResults.rds 
+#   data/processed/metaCalibrateOverall.rds
+# Output:
+#   figures/plotMeta.pdf
+
+library(tidyverse)
+
+calibrateOverallResults <- readRDS("data/processed/calibrateOverallResults.rds")
+metaCalibrateOverall    <- readRDS("data/processed/metaCalibrateOverall.rds")
+
+overall <- calibrateOverallResults %>%
+    select(-contains("Rr")) %>%
+    mutate(
+        type = "single",
+        position = case_when(
+            database == "mdcr" ~ 5,
+            database == "optum_dod" ~ 4,
+            database == "panther" ~ 3
+        )
+    )
+
+metaOverall <- metaCalibrateOverall %>%
+    mutate(
+        database = "overall",
+        type     = "meta",
+        position = 1
+    )
+
+combined <- rbind(overall, metaOverall) %>%
+    mutate(
+        database = factor(
+            x = database,
+            levels = c("overall", "optum_dod", "panther", "mdcr"),
+            labels = c("Overall", "OPTUM (DoD)", "Panther", "MDCR")
+        ),
+        outcome = factor(
+            x = outcome,
+            levels = 101:103,
+            labels = c(
+                "Hip fracture",
+                "Major Osteoporotic fracture",
+                "Vertebral fracture"
+            )
+        )
+    )
+
+ggplot(
+    data = combined,
+    aes(
+        x    = hr,
+        y    = position,
+        xmin = lower,
+        xmax = upper,
+        color = type,
+    )
+) +
+    facet_wrap(
+        ~outcome
+    ) +
+    geom_point(
+        aes(
+            shape = type, 
+            fill  = type
+        ), 
+        size = 2
+    ) +
+    geom_errorbar(width = 0) +
+    geom_vline(xintercept = 1, linetype = 2) +
+    scale_y_continuous(
+        breaks = c(1, 3, 4, 5),
+        labels = c(
+            "Summary",
+            "Optum EHR",
+            "Optum (DoD)",
+            "MDCR"
+        )
+    ) +
+    scale_color_manual(
+        breaks = c("meta", "single"),
+        values = c("red", "black")
+    ) +
+    scale_fill_manual(
[…]

### 2021-02-27 submission/manuscript.rmd
add figure to plot and format numbers

@@
 header-includes:
-  - \usepackage{helvet}
   - \renewcommand*\familydefault{\sfdefault}
@@
   - \usepackage[left]{lineno}
-  - \linenumbers
 editor_options: 
@@
 library(tidyverse)
+library(pander)
 library(knitr)
-options(scipen = 1, digits = 2)
+
+d <- function(x, decimals = 2) {
+  sprintf(paste0("%1.", decimals, "f"), x) 
+}
+
 knit_hooks$set(
@@
 
+\vspace{35mm}
+
+Alexandros Rekkas${^1}$, Annika M. J&ouml;dicke${^2}$, David van Klaveren$^3$, Daniel Prieto-Alhambra${^{1,2}}$, Peter R. Rijnbeek${^1}$
+
+\vspace{40mm}
+
+$^1$ Department of Medical Informatics, Erasmus University Medical Center, Rotterdam, Netherlands
+
+$^2$ Pharmaco- and Device Epidemiology, Centre for Statistics in Medicine, Nuffield Department of Orthopaedics, Rheumatology and Musculoskeletal Sciences, University of Oxford, Oxford, UK
+
+$^3$ Department of Public Health, Erasmus University Medical Center, Rotterdam, Netherlands
+
 \newpage
+\linenumbers
 # Introduction
@@
 ##     Statistical analyses
-We performed two sets of analyses.
+We carried out two sets of analyses.
 First, we derived overall treatment effect estimates of teriparatide compared to oral bisphosphonates regarding the three outcomes of interest.
@@
   here::here("data/processed/metaCalibrateOverall.rds")
+) %>%
+  mutate(
+    hr = d(hr),
+    lower = d(lower),
+    upper = d(upper)
   )
@@
 
-The random effects meta-analysis of teriparatide compared to oral bisphosphonates across the 3 considered databases gave a calibrated overall hazard ratio of `r metaCalibrateOverall %>% filter(outcome == 101) %>% pull(hr)` (`r metaCalibrateOverall %>% filter(outcome == 101) %>% pull(lower)` to `r metaCalibrateOverall %>% filter(outcome == 101) %>% pull(upper)`; 95% CI) for hip fracture, `r metaCalibrateOverall %>% filter(outcome == 102) %>% pull(hr)` (`r metaCalibrateOverall %>% filter(outcome == 102) %>% pull(lower)` to `r metaCalibrateOverall %>% filter(outcome == 102) %>% pull(upper)`; 95%CI) for major osteoporotic fracture and `r metaCalibrateOverall %>% filter(outcome == 103) %>% pull(hr)` (`r metaCalibrateOverall %>% filter(outcome == 103) %>% pull(lower)` to `r metaCalibrateOverall %>% filter(outcome == 103) %>% pull(upper)`; 95%CI) for vertebral fracture (##Figures 1-3).
+The random effects meta-analysis of teriparatide compared to oral bisphosphonates across the 3 considered databases gave a calibrated overall hazard ratio of `r metaCalibrateOverall %>% filter(outcome == 101) %>% pull(hr)` (`r metaCalibrateOverall %>% filter(outcome == 101) %>% pull(lower)` to `r metaCalibrateOverall %>% filter(outcome == 101) %>% pull(upper)`; 95% CI) for hip fracture, `r metaCalibrateOverall %>% filter(outcome == 102) %>% pull(hr)` (`r metaCalibrateOverall %>% filter(outcome == 102) %>% pull(lower)` to `r metaCalibrateOverall %>% filter(outcome == 102) %>% pull(upper)`; 95%CI) for major osteoporotic fracture and `r metaCalibrateOverall %>% filter(outcome == 103) %>% pull(hr)` (`r metaCalibrateOverall %>% filter(outcome == 103) %>% pull(lower)` to `r metaCalibrateOverall %>% filter(outcome == 103) %>% pull(upper)`; 95%CI) for vertebral fracture (Figure \@ref(fig:plotMeta)).
 
@@
   here::here("data/processed/metaCalibrateRiskStratified.rds")
+) %>%
+  mutate(
+    hr = d(hr),
+    lower = d(lower),
+    upper = d(upper)
   )
@@
   mutate(
-    hr = exp(logRr),
-    lower = exp(logLb95Rr),
-    upper = exp(logUb95Rr)
+    hr = d(exp(logRr)),
+    lower = d(exp(logLb95Rr)),
+    upper = d(exp(logUb95Rr))
   )
@@
   mutate(
-    estimate = 100 * estimate,
-    lower    = 100 * lower,
-    upper    = 100 * upper
+    estimate = d(100 * estimate),
+    lower    = d(100 * lower),
+    upper    = d(100 * upper)
   )
@@
 -->
+
+\newpage
+```{r plotMeta, echo=FALSE, fig.cap="This is a caption", out.width = '100%'}
+knitr::include_graphics(here::here("figures/plotMeta.pdf"))
+```

### 2021-02-27 code/plotMetaOverall.R
add result direction to figure

@@ -72,6 +72,36 @@ ggplot(
     ) +
     geom_errorbar(width = 0) +
     geom_vline(xintercept = 1, linetype = 2) +
+    # geom_segment(
+    #     x     = .5, 
+    #     xend  = .95, 
+    #     y     = 0, 
+    #     yend  = 0,
+    #     arrow = arrow(ends = "first", length = unit(2, "mm")),
+    #     color = "black"
+    # ) +
+    geom_text(
+        label = "Favors\nTeriparatide", 
+        x     = .73, 
+        y     = -.1, 
+        color = "black",
+        size  = 2
+    ) +
+    # geom_segment(
+    #     x     = 1.05, 
+    #     xend  = 1.5, 
+    #     y     = 0, 
+    #     yend  = 0,
+    #     arrow = arrow(ends = "last", length = unit(2, "mm")),
+    #     color = "black"
+    # ) +
+    geom_text(
+        label = "Favors\nBisphosphonates", 
+        x     = 1.3, 
+        y     = -.1, 
+        color = "black",
+        size  = 2
+    ) +
     scale_y_continuous(
         breaks = c(1, 3, 4, 5),
         labels = c(
@@ -79,7 +109,13 @@ ggplot(
             "Optum EHR",
             "Optum (DoD)",
             "MDCR"
-        )
+        ),
+        limits = c(-.2, 5)
+    ) +
+    scale_x_continuous(
+        breaks = c(.5, 1, 1.5),
+        labels = c("0.5", "1", "1.5"),
+        limits = c(.5, 1.5)
     ) +
     scale_color_manual(
         breaks = c("meta", "single"),
@@ -101,5 +137,5 @@ ggplot(
         strip.background = element_blank()
     )
 
-ggsave("figures/plotMeta.pdf", height = 2.95, width = 6)
+ggsave("figures/plotMeta.pdf", height = 3, width = 7)
 

### 2021-02-27 submission/manuscript.rmd
fix typos and rebuild

@@
 ## Study design
-We ran our analyses on two US observational administrative claims databases and one electronic health record (EHR) database, all mapped to OMOP-CDM version.
-More specifically, the considered claims databases were IBM MarketScan Medicare Supplemental Database (MDCR) representing retirees in the United States with primary or Medicare supplemental coverage and Optum De-Identified Clinformatics® Data Mart Database – Date of Death (DOD), including members of private health insurance.
-The EHR database was Optum de-identified Electronic Health Record Dataset (PANTHER) including Humedica’s Electronic Health Record.
+We ran our analyses on three US observational databases mapped to OMOP-CDM version. More specifically:
+
+- **IBM MarketScan Medicare Supplemental Database (MDCR)** is a claims database representing retirees in the United States with primary or Medicare supplemental coverage
+- **Optum De-Identified Clinformatics® Data Mart Database – Date of Death (DOD)** is a claims database including members of private health insurance.
+
+- **Optum de-identified Electronic Health Record Dataset (Optum-EHR)** is an EHR database including Humedica’s Electronic Health Record.
 
@@
 Patients were considered to be new-users if they received no anti-osteoporosis drugs (raloxifene, bazedoxifene, denosumab, abaloparatide, romosozumab) 365 days prior to treatment initiation with teriparatide or an oral bisphosphonate.
-We required patients to have at least one year of observation period before first exposure and an established osteoporosis diagnosis, defined as any recorded fracture of the hip, wrist, spine or shoulder/humerus.
+We required patients to have at least one year of observation period before first exposure and an established osteoporosis diagnosis, defined as any recorded fracture of the hip, wrist, spine or shoulder/humerus in their recorded medical history.
 
@@
 We excluded patients that had experienced the outcome under study any time prior to treatment initiation.
-Patient time-at-risk started at 1 day after treatment initiation until 730 days after treatment initiation.
+Patient time-at-risk started 1 day after treatment initiation and finished 730 days after treatment initiation.
 
@@
 We estimated treatment effects using Cox proportional hazards models with treatment as the sole covariate fitted in the 1-4 propensity score-matched subset of the considered patient population.
-Results from different databases were summarized using random effects meta-analysis (##REFERECNE).
+Results from different databases were summarized using random effects meta-analysis.
 
@@
 We considered the same large set of candidate covariates for the development of our prediction models.
-Time at risk was set at 1 to 730 days after study inclusion.
 
-For each outcome we used the derived prediction models to divide th e population under study in 4 equally-sized subgroups, based on the quarters of the predicted risk distribution.
+For each outcome we used the derived prediction models to divide the population under study in 4 equally-sized subgroups, based on the quarters of the predicted risk distribution.
 Within each of these risk-based subgroups we developed a new propensity score model.
@@
 We derived relative effect estimates using Cox proportional hazards models only with treatment as a predictor.
-Absolute effect estimates were calculated based on the difference of the Kaplan-Meier estimates.
+Absolute effect estimates were calculated based on the difference of the Kaplan-Meier estimates, on day 730 after treatment initiation.
 
@@
 
-When assessing heterogeneity of treatment effect, we estimated calibrated hazard ratios of `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "mdcr") %>% pull(hr)` (`r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "mdcr") %>% pull(lower)` to `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "mdcr") %>% pull(upper)`; 95% CI),
-`r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "optum_dod") %>% pull(hr)` (`r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "optum_dod") %>% pull(lower)` to `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "optum_dod") %>% pull(upper)`; 95% CI) and `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "panther") %>% pull(hr)` (`r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "panther") %>% pull(hr)` to `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "panther") %>% pull(hr)`; 95% CI) within the highest risk quarter of MDCR, OPTUM_DOD and PANTHER respectively.
+When assessing heterogeneity of treatment effect for hip fractur    e, we estimated calibrated hazard ratios of `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "mdcr") %>% pull(hr)` (`r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "mdcr") %>% pull(lower)` to `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "mdcr") %>% pull(upper)`; 95% CI),
+`r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "optum_dod") %>% pull(hr)` (`r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "optum_dod") %>% pull(lower)` to `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "optum_dod") %>% pull(upper)`; 95% CI) and `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "panther") %>% pull(hr)` (`r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "panther") %>% pull(lower)` to `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "panther") %>% pull(upper)`; 95% CI) within the highest risk quarter of MDCR, OPTUM_DOD and PANTHER respectively.
 This translated in absolute risk reduction of `r hipFractureAbsolute %>% filter(database == "mdcr") %>% pull(estimate)`% (`r hipFractureAbsolute %>% filter(database == "mdcr") %>% pull(lower)`% to `r hipFractureAbsolute %>% filter(database == "mdcr") %>% pull(upper)`%; 95% CI), `r hipFractureAbsolute %>% filter(database == "optum_dod") %>% pull(estimate)`% (`r hipFractureAbsolute %>% filter(database == "optum_dod") %>% pull(lower)`% to `r hipFractureAbsolute %>% filter(database == "optum_dod") %>% pull(upper)`%; 95% CI), `r hipFractureAbsolute %>% filter(database == "panther") %>% pull(estimate)`% (`r hipFractureAbsolute %>% filter(database == "panther") %>% pull(lower)`% to `r hipFractureAbsolute %>% filter(database == "panther") %>% pull(upper)`%; 95% CI).

### 2021-02-27 code/plotMetaOverall.R

@@ -44,7 +44,7 @@ combined <- rbind(overall, metaOverall) %>%
             levels = 101:103,
             labels = c(
                 "Hip fracture",
-                "Major Osteoporotic fracture",
+                "Major osteoporotic fracture",
                 "Vertebral fracture"
             )
         )
@@ -72,14 +72,6 @@ ggplot(
     ) +
     geom_errorbar(width = 0) +
     geom_vline(xintercept = 1, linetype = 2) +
-    # geom_segment(
-    #     x     = .5, 
-    #     xend  = .95, 
-    #     y     = 0, 
-    #     yend  = 0,
-    #     arrow = arrow(ends = "first", length = unit(2, "mm")),
-    #     color = "black"
-    # ) +
     geom_text(
         label = "Favors\nTeriparatide", 
         x     = .73, 
@@ -87,14 +79,6 @@ ggplot(
         color = "black",
         size  = 2
     ) +
-    # geom_segment(
-    #     x     = 1.05, 
-    #     xend  = 1.5, 
-    #     y     = 0, 
-    #     yend  = 0,
-    #     arrow = arrow(ends = "last", length = unit(2, "mm")),
-    #     color = "black"
-    # ) +
     geom_text(
         label = "Favors\nBisphosphonates", 
         x     = 1.3, 

### 2021-02-28 submission/manuscript.rmd
introduction citations

@@
   chunk_output_type: console
+bibliography: references.bib
+csl: jamia.csl
 ---
@@
 # Introduction
-Osteoporosis is a common condition characterised by decreased bone density and associated with increased risk for fragility fractures, which affects almost 30% of women aged $\geq50$ years (#Wright et al.
-2017).
+Osteoporosis is a common condition characterised by decreased bone density and associated with increased risk for fragility fractures, which affects almost 30% of women aged $\geq50$ years [@wright2017impact].
 It was estimated that in 2005, over 2 million incident fractures occurred in the United States, with hip fractures accounting for more than 70% of the costs.
-By 2025, annual fractures and costs could rise by almost 50% (#Cole et al.
-2008).
+By 2025, annual fractures and costs could rise by almost 50% [@Cole2008].
 
@@
 Teriparatide, a parathyroid hormone analogue administered as a daily injection, was approved by the FDA in 2002 as the first anabolic agent for treatment of severe postmenopausal osteoporosis.
-While clinical trials among patients with previous vertebral fractures showed a substantial risk reduction for new vertebral fracture comparing teriparatide to placebo (#Neer et al 2001) and risedronate (#Kendler et al.
-2018), its efficacy on low-incident major osteoporotic fractures, especially hip fractures, is less well established: Previous randomised trials assessing hip fracture comprised only few events, thus not providing sufficient power to show differences between treatment groups when studying comparative effectiveness.
-A recent meta-analysis by Diez-Perez and colleagues assessed the effect of teriparatide on hip fracture, indicating a significant 80% risk reduction compared to placebo and a non-significant 46% risk reduction when compared to active controls.
+While clinical trials among patients with previous vertebral fractures showed a substantial risk reduction for new vertebral fracture comparing teriparatide to placebo [@Neer2001] and risedronate [@Kendler2018], its efficacy on low-incident major osteoporotic fractures, especially hip fractures, is less well established: Previous randomised trials assessing hip fracture comprised only few events, thus not providing sufficient power to show differences between treatment groups when studying comparative effectiveness.
+A recent meta-analysis by Diez-Perez and colleagues assessed the effect of teriparatide on hip fracture, indicating a significant 80% risk reduction compared to placebo and a non-significant 46% risk reduction when compared to active controls [@DezPrez2019].
 
@@
 While hip fracture is a comparatively rare outcome, it is unambiguously defined and reliably recorded in routinely collected data, allowing for a direct comparison to results from clinical trial meta-analyses.
-While results from non-controlled, observational studies suggest that teriparatide may reduce the risk of hip fractures (#Burge et al.
-2017, #Silverman et al.
-2019), this is the first observational study assessing teriparatide in hip fracture risk reduction compared to oral bisphosphonate users in postmenopausal women using multiple large real-world databases.
+While results from non-controlled, observational studies suggest that teriparatide may reduce the risk of hip fractures [@Burge2016; @Silverman2018], this is the first observational study assessing teriparatide in hip fracture risk reduction compared to oral bisphosphonate users in postmenopausal women using multiple large real-world databases.
 
@@
 \newpage
+# Discussion
+
+\newpage
+# References
+\setlength{\parindent}{-0.25in}
+\setlength{\leftskip}{0.25in}
+\noindent
+
+<div id="refs"></div>
+
+\setlength{\parindent}{0in}
+\setlength{\leftskip}{0in}
+\noindent
+
+\newpage
 ```{r plotMeta, echo=FALSE, fig.cap="This is a caption", out.width = '100%'}

### 2021-02-28 code/plotMetaOverall.R

@@ -50,7 +50,7 @@ combined <- rbind(overall, metaOverall) %>%
         )
     )
 
-ggplot(
+p <- ggplot(
     data = combined,
     aes(
         x    = hr,
@@ -121,5 +121,5 @@ ggplot(
         strip.background = element_blank()
     )
 
-ggsave("figures/plotMeta.pdf", height = 3, width = 7)
+ggsave("figures/plotMeta.pdf", plot = p, height = 3, width = 7)
 

### 2021-02-28 code/plotMetaOverall.R
include tiff figure

@@ -122,4 +122,5 @@ p <- ggplot(
     )
 
 ggsave("figures/plotMeta.pdf", plot = p, height = 3, width = 7)
+ggsave("figures/plotMeta.tiff", plot = p, height = 3, width = 7, compression = "lzw+p")
 

### 2021-02-28 submission/manuscript.rmd
include tiff figure
planning absolute plot

@@
 
-When assessing heterogeneity of treatment effect for hip fractur    e, we estimated calibrated hazard ratios of `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "mdcr") %>% pull(hr)` (`r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "mdcr") %>% pull(lower)` to `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "mdcr") %>% pull(upper)`; 95% CI),
+When assessing heterogeneity of treatment effect for hip fracture, we estimated calibrated hazard ratios of `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "mdcr") %>% pull(hr)` (`r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "mdcr") %>% pull(lower)` to `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "mdcr") %>% pull(upper)`; 95% CI),
 `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "optum_dod") %>% pull(hr)` (`r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "optum_dod") %>% pull(lower)` to `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "optum_dod") %>% pull(upper)`; 95% CI) and `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "panther") %>% pull(hr)` (`r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "panther") %>% pull(lower)` to `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "panther") %>% pull(upper)`; 95% CI) within the highest risk quarter of MDCR, OPTUM_DOD and PANTHER respectively.
@@
 ```{r plotMeta, echo=FALSE, fig.cap="This is a caption", out.width = '100%'}
-knitr::include_graphics(here::here("figures/plotMeta.pdf"))
+knitr::include_graphics(here::here("figures/plotMeta.tiff"))
 ```

### 2021-02-28 code/plotAbsoluteHip.R

@@ -1,0 +1,104 @@
+# Add mean predicted risk, and rectangle limits to absolute dataframe
+
+
+library(tidyverse)
+
+absolute <- readRDS("data/raw/mappedOverallAbsoluteResults.rds") %>%
+    filter(
+        analysisType == "matchOnPs_1_to_4",
+        stratOutcome == 101,
+        estOutcome == 101
+    ) %>%
+    mutate(
+        estimate = 100 * estimate,
+        lower    = 100 * lower,
+        upper    = 100 * upper,
+        stratum = case_when(
+            riskStratum == "Q1" ~1,
+            riskStratum == "Q2" ~2,
+            riskStratum == "Q3" ~3,
+            riskStratum == "Q4" ~4
+        )
+    )
+
+dataRect <- function(xmin, xmax) {
+    data.frame(
+        xmin = xmin,
+        xmax = xmax,
+        ymin = -Inf,
+        ymax = Inf
+    )
+}
+
+ggplot() +
+    geom_rect(
+        data = dataRect(-Inf, 1.5),
+        inherit.aes = FALSE,
+        aes(
+            xmin = xmin,
+            xmax = xmax,
+            ymin = ymin,
+            ymax = ymax
+        ),
+        fill = "#ffffcc",
+        alpha = .3
+    ) +
+    geom_rect(
+        data = dataRect(1.5, 2.5),
+        inherit.aes = F,
+        aes(
+            xmin = xmin,
+            xmax = xmax,
+            ymin = ymin,
+            ymax = ymax
+        ),
+        fill = "#a1dab4",
+        alpha = .3
+    ) +
+    geom_rect(
+        data = dataRect(2.5, 3.5),
+        inherit.aes = F,
+        aes(
+            xmin = xmin,
+            xmax = xmax,
+            ymin = ymin,
+            ymax = ymax
+        ),
+        fill = "#41b6c4",
+        alpha = .3
+    ) +
+    geom_rect(
+        data = dataRect(3.5, Inf),
+        inherit.aes = F,
+        aes(
+            xmin = xmin,
+            xmax = xmax,
+            ymin = ymin,
+            ymax = ymax
+        ),
+        fill = "#225ea8",
+        alpha = .3
+    ) +
+    geom_point(
+        data = absolute,
+        aes(
+            x = stratum,           # This will be the mean predicted risk
+            y = estimate
+        )
+        
+    ) +
+    geom_errorbar(
+        data = absolute,
[…]

### 2021-03-01 submission/manuscript.rmd
incorporate risk stratified plot for hip fracture

@@
 library(tidyverse)
-library(pander)
 library(knitr)
@@
 ```
+
+\newpage
+```{r plotAbsoluteHip, echo=FALSE, fig.cap="This is a caption", out.width = '100%'}
+knitr::include_graphics(here::here("figures/plotAbsoluteHip.png"))
+```

### 2021-03-01 code/plotAbsoluteHip.R

@@ -1,5 +1,19 @@
-# Add mean predicted risk, and rectangle limits to absolute dataframe
+#!/usr/bin/env Rscript
 
+# Author: Alexandros Rekkas
+#
+# Description:
+#   Plots the risk stratified absolute difference for hip fracture with respect
+#   to estimated hip fracture risk
+# Depends:
+#   data/raw/mappedOverallAbsoluteResults.rds
+# Output:
+#   figures/plotAbsoluteHip.pdf
+#   figures/plotAbsoluteHip.tiff
+#   figures/plotAbsoluteHip.png
+# 
+# Notes:
+#   Add mean predicted risk, and rectangle limits to absolute dataframe
 
 library(tidyverse)
 
@@ -13,11 +27,16 @@ absolute <- readRDS("data/raw/mappedOverallAbsoluteResults.rds") %>%
         estimate = 100 * estimate,
         lower    = 100 * lower,
         upper    = 100 * upper,
-        stratum = case_when(
+        meanRisk = case_when(
             riskStratum == "Q1" ~1,
             riskStratum == "Q2" ~2,
             riskStratum == "Q3" ~3,
             riskStratum == "Q4" ~4
+        ),
+        database = case_when(
+            database == "mdcr" ~ "MDCR",
+            database == "optum_dod" ~ "Optum-DOD",
+            database == "panther" ~ "Optum-EHR"
         )
     )
 
@@ -30,7 +49,7 @@ dataRect <- function(xmin, xmax) {
     )
 }
 
-ggplot() +
+p <- ggplot() +
     geom_rect(
         data = dataRect(-Inf, 1.5),
         inherit.aes = FALSE,
@@ -82,7 +101,7 @@ ggplot() +
     geom_point(
         data = absolute,
         aes(
-            x = stratum,           # This will be the mean predicted risk
+            x = meanRisk,           # This will be the mean predicted risk
             y = estimate
         )
         
@@ -92,13 +111,32 @@ ggplot() +
         aes(
             ymin = lower, 
             ymax = upper, 
-            x = stratum, 
+            x = meanRisk, 
             y = estimate
         ), 
         width = 0
     ) +
     geom_hline(yintercept = 0, linetype = "dashed") +
-    facet_wrap(~database, ncol = 1) +
-    ylim(-3, 5) +
-    theme_classic()
+    facet_wrap(
+        ~database, 
+        ncol = 1,
+        strip.position = "left"
[…]

### 2021-03-01 code/plotMetaOverall.R

@@ -123,4 +123,5 @@ p <- ggplot(
 
 ggsave("figures/plotMeta.pdf", plot = p, height = 3, width = 7)
 ggsave("figures/plotMeta.tiff", plot = p, height = 3, width = 7, compression = "lzw+p")
+ggsave("figures/plotMeta.png", plot = p, height = 3, width = 7)
 

### 2021-03-01 extras/init.R

@@ -1,0 +1,4 @@
+#!/usr/bin/env Rscript
+
+install.packages("remotes")
+remotes::install_deps(dependencies = TRUE)

### 2021-03-01 extras/init.sh

@@ -1,0 +1,4 @@
+#!/bin/bash
+
+mkdir -p data/raw data/processed
+./init.R

### 2021-03-01 submission/manuscript.rmd
add more references

--- 2021-02-28 submission/manuscript.rmd
@@
 library(tidyverse)
-library(pander)
 library(knitr)
+library(kableExtra)
 
@@
 - **IBM MarketScan Medicare Supplemental Database (MDCR)** is a claims database representing retirees in the United States with primary or Medicare supplemental coverage
-- **Optum De-Identified Clinformatics® Data Mart Database – Date of Death (DOD)** is a claims database including members of private health insurance.
+- **Optum De-Identified Clinformatics® Data Mart Database – Date of Death (Optum-DOD)** is a claims database including members of private health insurance.
 
@@
 
-Our cohorts included female new users of teriparatide or oral bisphosphonates above the age of 65 (ref 7,8 LEGEND-HTN).
+Our cohorts included female new users of teriparatide or oral bisphosphonates above the age of 65 [@Ryan2013].
 Patients were considered to be new-users if they received no anti-osteoporosis drugs (raloxifene, bazedoxifene, denosumab, abaloparatide, romosozumab) 365 days prior to treatment initiation with teriparatide or an oral bisphosphonate.
@@
 First, we derived overall treatment effect estimates of teriparatide compared to oral bisphosphonates regarding the three outcomes of interest.
-To account for potential measured confounding we developed separate propensity score models within each database using the same predefined set of measured covariates.
-Covariates were included in the propensity score model using regularized logistic regression (##REFERENCE Tian-Schuemie simulation).
-We estimated treatment effects using Cox proportional hazards models with treatment as the sole covariate fitted in the 1-4 propensity score-matched subset of the considered patient population.
+To account for potential measured confounding we developed separate large-scale propensity score models within each database based on LASSO logistic regression using the same predefined set of measured covariates [@Tian2018].
+We estimated treatment effects using Cox proportional hazards models with treatment as the sole covariate fitted in the 1:4 propensity score-matched subset of the considered patient population.
 Results from different databases were summarized using random effects meta-analysis.
 
-For the second set of analyses, we used the standardized framework for risk-based assessment of treatment effect heterogeneity, that follows the directives of the PATH statement (##REFERENCE framework).
+For the second set of analyses, we used the standardized framework for risk-based assessment of treatment effect heterogeneity, that follows the directives of the PATH statement [@Kent2019; @KentEnE2019].
 Following data extraction, the framework was applied in two steps.
 First, we derived individualized risk predictions for the three efficacy outcomes.
-We built the prediction models using LASSO logistic regression on the propensity score matched (1 to 4) subpopulation of the pooled treatment arms, aiming to remove any effect of measured confounding that may cause differential fit of our models across treatment arms.
+We built the prediction models using LASSO logistic regression on the propensity score matched (1:4) subpopulation of the pooled treatment arms, aiming to remove any effect of measured confounding that may cause differential fit of our models across treatment arms.
 We considered the same large set of candidate covariates for the development of our prediction models.
@@
 Within each of these risk-based subgroups we developed a new propensity score model.
-Our analyses were performed on the propensity score matched (1 to 4) subset of the risk subgroup subset.
-We used a set of XXX negative control outcomes, to calibrate our results for unmeasured confounding (##REFEFRENCE).
+Our analyses were performed on the propensity score matched (1:4) subset of the risk subgroup subset.
 We derived relative effect estimates using Cox proportional hazards models only with treatment as a predictor.
 Absolute effect estimates were calculated based on the difference of the Kaplan-Meier estimates, on day 730 after treatment initiation.
+
+Residual study bias from unmeasured confounding can still be present in observational studies, which often may not be visible when evaluating propensity score adjustment.
+To account for that, we considered a set of 126 negative control outcome experiments, where a null effect was assumed to be true.
+We used the estimated relative effects for these outcomes to derive an empirical approximation to the true null distribution, which was then used to calibrate the hazard ratios and their 95% confidence intervals for the 3 outcomes of interest.
 
@@
 When assessing heterogeneity of treatment effect for hip fracture, we estimated calibrated hazard ratios of `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "mdcr") %>% pull(hr)` (`r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "mdcr") %>% pull(lower)` to `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "mdcr") %>% pull(upper)`; 95% CI),
-`r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "optum_dod") %>% pull(hr)` (`r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "optum_dod") %>% pull(lower)` to `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "optum_dod") %>% pull(upper)`; 95% CI) and `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "panther") %>% pull(hr)` (`r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "panther") %>% pull(lower)` to `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "panther") %>% pull(upper)`; 95% CI) within the highest risk quarter of MDCR, OPTUM_DOD and PANTHER respectively.
-This translated in absolute risk reduction of `r hipFractureAbsolute %>% filter(database == "mdcr") %>% pull(estimate)`% (`r hipFractureAbsolute %>% filter(database == "mdcr") %>% pull(lower)`% to `r hipFractureAbsolute %>% filter(database == "mdcr") %>% pull(upper)`%; 95% CI), `r hipFractureAbsolute %>% filter(database == "optum_dod") %>% pull(estimate)`% (`r hipFractureAbsolute %>% filter(database == "optum_dod") %>% pull(lower)`% to `r hipFractureAbsolute %>% filter(database == "optum_dod") %>% pull(upper)`%; 95% CI), `r hipFractureAbsolute %>% filter(database == "panther") %>% pull(estimate)`% (`r hipFractureAbsolute %>% filter(database == "panther") %>% pull(lower)`% to `r hipFractureAbsolute %>% filter(database == "panther") %>% pull(upper)`%; 95% CI).
+`r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "optum_dod") %>% pull(hr)` (`r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "optum_dod") %>% pull(lower)` to `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "optum_dod") %>% pull(upper)`; 95% CI) and `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "panther") %>% pull(hr)` (`r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "panther") %>% pull(lower)` to `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "panther") %>% pull(upper)`; 95% CI) within the highest risk quarter of MDCR, Optum-DOD and Optum-EHR respectively.
+This translated in absolute risk reduction of `r hipFractureAbsolute %>% filter(database == "mdcr") %>% pull(estimate)`% (`r hipFractureAbsolute %>% filter(database == "mdcr") %>% pull(lower)`% to `r hipFractureAbsolute %>% filter(database == "mdcr") %>% pull(upper)`%; 95% CI), `r hipFractureAbsolute %>% filter(database == "optum_dod") %>% pull(estimate)`% (`r hipFractureAbsolute %>% filter(database == "optum_dod") %>% pull(lower)`% to `r hipFractureAbsolute %>% filter(database == "optum_dod") %>% pull(upper)`%; 95% CI), `r hipFractureAbsolute %>% filter(database == "panther") %>% pull(estimate)`% (`r hipFractureAbsolute %>% filter(database == "panther") %>% pull(lower)`% to `r hipFractureAbsolute %>% filter(database == "panther") %>% pull(upper)`%; 95% CI) respectively (Figure \@ref(fig:plotAbsoluteHip)).
 The random effects meta-analytic calibrated hazard ratio within the highest risk quarter was `r metaCalibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101) %>% pull(hr)` (`r metaCalibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101) %>% pull(lower)` to `r metaCalibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101) %>% pull(upper)`; 95% CI) across the 3 databases.
+The full set of results can be explored at https://arekkas.shinyapps.io/ter_bis_3dbs/.
 
@@
 \noindent
-
 <div id="refs"></div>
-
 \setlength{\parindent}{0in}
@@
 \newpage
-```{r plotMeta, echo=FALSE, fig.cap="This is a caption", out.width = '100%'}
-knitr::include_graphics(here::here("figures/plotMeta.tiff"))
+# Tables and figures
+```{r, echo=FALSE}
+incidenceOverall <- readRDS(here::here("data/raw/incidenceOverall.rds"))
+incidence <- incidenceOverall %>% 
+  filter(analysisType == "matchOnPs_1_to_4") %>%
+  mutate(
+    database = case_when(
+      database == "mdcr" ~ "MDCR",
+      database == "optum_dod" ~ "Optum-DOD",
+      database == "panther" ~ "Optum-EHR"
+    ),
+    treatmentDays = round(treatmentDays / 365),
+    comparatorDays = round(comparatorDays / 365)
+  ) %>%
+  select(
+    database,
+    outcomeId,
+    treatmentPersons,
+    treatmentDays,
+    treatmentOutcomes,
+    comparatorPersons,
+    comparatorDays,
+    comparatorOutcomes
+  ) %>%
+  arrange(database, outcomeId) %>%
+  mutate(
+    database = ifelse(
+      outcomeId == 101,
+      database,
+      ""
+    ),
+    outcomeId = case_when(
+      outcomeId == 101 ~ "Hip fracture",
+      outcomeId == 102 ~ "Major osteoporotic\nfracture",
+      outcomeId == 103 ~ "Vertebral fracture"
+    ),
+  )
+
+kable(
+  incidence,
+  longtable = TRUE,
+  booktabs = TRUE,
+  col.names = c(
+    "Database",
+    "Outcome",
+    "Persons",
+    "Person years",
+    "Events",
+    "Persons",
+    "Person years",
+    "Events"
+  ),
+  format.args = list(
+    big.mar = ","
+  )
+) %>%
+  add_header_above(c("", "", "Teriparatide" = 3, "Bisphosphonates" = 3))
+  # kableExtra::add_indent(c(1, 4, 7))
 ```
+
+\newpage
+```{r plotMeta, echo=FALSE, fig.cap="Calibrated hazard ratios for the 3 outcomes of interest across the 3 considered databases using a set of 126 negative controls.
+Values below 1 favour teriparatide, while values above 1 favour bisphosphonates.", out.width = '100%'}
+knitr::include_graphics(here::here("figures/plotMeta.png"))
+```
+
+\newpage
+```{r plotAbsoluteHip, echo=FALSE, fig.cap="Absolute risk differences wihtin quarters of predicted hip fracture risk across the 3 considered databases.", out.width = '100%'}
+knitr::include_graphics(here::here("figures/plotAbsoluteHip.png"))
+```

### 2021-03-01 code/plotAbsoluteHip.R

@@ -21,7 +21,7 @@ absolute <- readRDS("data/raw/mappedOverallAbsoluteResults.rds") %>%
     filter(
         analysisType == "matchOnPs_1_to_4",
         stratOutcome == 101,
-        estOutcome == 101
+        estOutcome   == 101
     ) %>%
     mutate(
         estimate = 100 * estimate,

### 2021-03-01 code/plotMetaOverall.R
fix labels in plot

@@ -90,8 +90,8 @@ p <- ggplot(
         breaks = c(1, 3, 4, 5),
         labels = c(
             "Summary",
-            "Optum EHR",
-            "Optum (DoD)",
+            "Optum-EHR",
+            "Optum-DOD",
             "MDCR"
         ),
         limits = c(-.2, 5)

### 2021-03-01 submission/manuscript.rmd
proof reading

@@
 \newpage
+
 \linenumbers
@@
 We built the prediction models using LASSO logistic regression on the propensity score matched (1:4) subpopulation of the pooled treatment arms, aiming to remove any effect of measured confounding that may cause differential fit of our models across treatment arms.
-We considered the same large set of candidate covariates for the development of our prediction models.
+We considered the same large set of candidate covariates as for the development of the propensity score models.
 
-For each outcome we used the derived prediction models to divide the population under study in 4 equally-sized subgroups, based on the quarters of the predicted risk distribution.
+For each outcome we used the derived prediction models to divide the population in 4 equally-sized subgroups, based on the quarters of the predicted risk distribution.
 Within each of these risk-based subgroups we developed a new propensity score model.
@@
 To account for that, we considered a set of 126 negative control outcome experiments, where a null effect was assumed to be true.
-We used the estimated relative effects for these outcomes to derive an empirical approximation to the true null distribution, which was then used to calibrate the hazard ratios and their 95% confidence intervals for the 3 outcomes of interest.
+We used the estimated relative effects for these outcomes to derive an empirical approximation to the true null distribution, which was then used to calibrate the hazard ratios and their 95% confidence intervals for the 3 outcomes of interest [@Schuemie2013; @Schuemie2016].
 
@@
 A total of `r sum(incidenceOverall$maxTreatment) + sum(incidenceOverall$maxComparator)` patients were initially considered for our analyses.
-Of those, `r sum(incidenceOverall$maxTreatment)` received teriparatide and `r sum(incidenceOverall$maxComparator)` received an oral bisphosphonate (##TABLE1).
+Of those, `r sum(incidenceOverall$maxTreatment)` received teriparatide and `r sum(incidenceOverall$maxComparator)` received an oral bisphosphonate \@ref(tab:table1).
 
@@
 
+hipRiskPerformance <- readRDS(here::here("data/raw/predictionPerformance.rds")) %>%
+  filter(
+    cohort == "EntirePopulation",
+    stratOutcome == 101
+  )
+
 ```
 
+The internally developed prediction models for hip fracture showed adequate performance achieving AUC of `r hipRiskPerformance %>% filter(database == "mdcr") %>% pull(auc) %>% d()`, `r hipRiskPerformance %>% filter(database == "optum_dod") %>% pull(auc) %>% d()` and `r hipRiskPerformance %>% filter(database == "panther") %>% pull(auc) %>% d()` in MDCR, Optum-DOD and Optum-EHR respectively.
 When assessing heterogeneity of treatment effect for hip fracture, we estimated calibrated hazard ratios of `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "mdcr") %>% pull(hr)` (`r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "mdcr") %>% pull(lower)` to `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "mdcr") %>% pull(upper)`; 95% CI),
@@
 The random effects meta-analytic calibrated hazard ratio within the highest risk quarter was `r metaCalibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101) %>% pull(hr)` (`r metaCalibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101) %>% pull(lower)` to `r metaCalibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101) %>% pull(upper)`; 95% CI) across the 3 databases.
+For the secondary outcomes (major osteoporotic fracture and vertebral fracture) we found no evidence of risk-based treatment effect heterogeneity.
+This was anticipated as overall effects of teriparatide on theses outcomes were not significant at the 95% confidence level.
 The full set of results can be explored at https://arekkas.shinyapps.io/ter_bis_3dbs/.
@@
 # Tables and figures
-```{r, echo=FALSE}
+```{r table1, echo=FALSE, results='asis'}
 incidenceOverall <- readRDS(here::here("data/raw/incidenceOverall.rds"))
@@
 
-kable(
+knitr::kable(
   incidence,
+  format   = "latex",
   longtable = TRUE,
@@
     big.mar = ","
-  )
+  ),
+   caption = "Number of patients, person years and outcomes for each outcome across all databases."
 ) %>%

### 2021-03-18 submission/manuscript.rmd
fix line width for readability

@@
 
-Alexandros Rekkas${^1}$, Annika M. J&ouml;dicke${^2}$, David van Klaveren$^3$, Daniel Prieto-Alhambra${^{1,2}}$, Peter R. Rijnbeek${^1}$
+Alexandros Rekkas${^1}$, Annika M. J&ouml;dicke${^2}$, David van Klaveren$^3$,
+Daniel Prieto-Alhambra${^{1,2}}$, Peter R. Rijnbeek${^1}$
 
@@
 
-$^1$ Department of Medical Informatics, Erasmus University Medical Center, Rotterdam, Netherlands
+$^1$ Department of Medical Informatics, Erasmus University Medical Center,
+Rotterdam, Netherlands
 
-$^2$ Pharmaco- and Device Epidemiology, Centre for Statistics in Medicine, Nuffield Department of Orthopaedics, Rheumatology and Musculoskeletal Sciences, University of Oxford, Oxford, UK
+$^2$ Pharmaco- and Device Epidemiology, Centre for Statistics in Medicine,
+Nuffield Department of Orthopaedics, Rheumatology and Musculoskeletal Sciences,
+University of Oxford, Oxford, UK
 
-$^3$ Department of Public Health, Erasmus University Medical Center, Rotterdam, Netherlands
+$^3$ Department of Public Health, Erasmus University Medical Center, Rotterdam,
+Netherlands
 
@@
 # Introduction
-Osteoporosis is a common condition characterised by decreased bone density and associated with increased risk for fragility fractures, which affects almost 30% of women aged $\geq50$ years [@wright2017impact].
-It was estimated that in 2005, over 2 million incident fractures occurred in the United States, with hip fractures accounting for more than 70% of the costs.
-By 2025, annual fractures and costs could rise by almost 50% [@Cole2008].
+Osteoporosis is a common condition characterised by decreased bone density and
+associated with increased risk for fragility fractures, which affects almost
+30% of women aged $\geq50$ years [@wright2017impact]. It was estimated that in
+2005, over 2 million incident fractures occurred in the United States, with hip
+fractures accounting for more than 70% of the costs. By 2025, annual fractures
+and costs could rise by almost 50% [@Cole2008].
 
-Fracture prevention is thus the key focus of anti-osteoporotic therapy, reducing the disease burden both for the affected patient but also on the population level.
-Several pharmacological agents are available for primary and secondary prevention.
-The choice of anti-osteoporotic agent largely depends on history of fragility fracture and anticipated fracture risk.
-Oral bisphosphonates (BP) are first line treatments for postmenopausal patients with increased fracture risk in clinical guidelines (#AACE/ACE 2020) as for their favourable cost-effectiveness and safety profile.
-Teriparatide, a parathyroid hormone analogue administered as a daily injection, was approved by the FDA in 2002 as the first anabolic agent for treatment of severe postmenopausal osteoporosis.
-While clinical trials among patients with previous vertebral fractures showed a substantial risk reduction for new vertebral fracture comparing teriparatide to placebo [@Neer2001] and risedronate [@Kendler2018], its efficacy on low-incident major osteoporotic fractures, especially hip fractures, is less well established: Previous randomised trials assessing hip fracture comprised only few events, thus not providing sufficient power to show differences between treatment groups when studying comparative effectiveness.
-A recent meta-analysis by Diez-Perez and colleagues assessed the effect of teriparatide on hip fracture, indicating a significant 80% risk reduction compared to placebo and a non-significant 46% risk reduction when compared to active controls [@DezPrez2019].
+Fracture prevention is thus the key focus of anti-osteoporotic therapy,
+reducing the disease burden both for the affected patient but also on the
+population level. Several pharmacological agents are available for primary and
+secondary prevention. The choice of anti-osteoporotic agent largely depends on
+history of fragility fracture and anticipated fracture risk. Oral
+bisphosphonates (BP) are first line treatments for postmenopausal patients with
+increased fracture risk in clinical guidelines (#AACE/ACE 2020) as for their
+favourable cost-effectiveness and safety profile. Teriparatide, a parathyroid
+hormone analogue administered as a daily injection, was approved by the FDA in
+2002 as the first anabolic agent for treatment of severe postmenopausal
+osteoporosis. While clinical trials among patients with previous vertebral
+fractures showed a substantial risk reduction for new vertebral fracture
+comparing teriparatide to placebo [@Neer2001] and risedronate [@Kendler2018],
+its efficacy on low-incident major osteoporotic fractures, especially hip
+fractures, is less well established: Previous randomised trials assessing hip
+fracture comprised only few events, thus not providing sufficient power to show
+differences between treatment groups when studying comparative effectiveness. A
+recent meta-analysis by Diez-Perez and colleagues assessed the effect of
+teriparatide on hip fracture, indicating a significant 80% risk reduction
+compared to placebo and a non-significant 46% risk reduction when compared to
+active controls [@DezPrez2019].
 
-With teriparatide-containing biosimilars being launched in recent years, treatment costs dropped significantly.
-Subsequently, the discussion of cost-effectiveness in patients with less severe forms of osteoporosis is restarting.
-Therefore, evaluation of the effectiveness in hip fracture prevention is required.
-In particular, observational studies assessing comparative effectiveness in a real-world setting are required, as effectiveness may differ from clinical trial as for differences in patient’s persistence and adherence to their anti-osteoporotic treatment.
-While hip fracture is a comparatively rare outcome, it is unambiguously defined and reliably recorded in routinely collected data, allowing for a direct comparison to results from clinical trial meta-analyses.
-While results from non-controlled, observational studies suggest that teriparatide may reduce the risk of hip fractures [@Burge2016; @Silverman2018], this is the first observational study assessing teriparatide in hip fracture risk reduction compared to oral bisphosphonate users in postmenopausal women using multiple large real-world databases.
+With teriparatide-containing biosimilars being launched in recent years,
+treatment costs dropped significantly. Subsequently, the discussion of
+cost-effectiveness in patients with less severe forms of osteoporosis is
+restarting. Therefore, evaluation of the effectiveness in hip fracture
+prevention is required. In particular, observational studies assessing
+comparative effectiveness in a real-world setting are required, as
+effectiveness may differ from clinical trial as for differences in patient’s
+persistence and adherence to their anti-osteoporotic treatment. While hip
+fracture is a comparatively rare outcome, it is unambiguously defined and
+reliably recorded in routinely collected data, allowing for a direct comparison
+to results from clinical trial meta-analyses. While results from
+non-controlled, observational studies suggest that teriparatide may reduce the
+risk of hip fractures [@Burge2016; @Silverman2018], this is the first
+observational study assessing teriparatide in hip fracture risk reduction
+compared to oral bisphosphonate users in postmenopausal women using multiple
+large real-world databases.
 
@@
 ## Study design
-We ran our analyses on three US observational databases mapped to OMOP-CDM version. More specifically:
+We ran our analyses on three US observational databases mapped to OMOP-CDM
+version. More specifically:
 
-- **IBM MarketScan Medicare Supplemental Database (MDCR)** is a claims database representing retirees in the United States with primary or Medicare supplemental coverage
-- **Optum De-Identified Clinformatics® Data Mart Database – Date of Death (Optum-DOD)** is a claims database including members of private health insurance.
+- **IBM MarketScan Medicare Supplemental Database (MDCR)** is a claims database
+  representing retirees in the United States with primary or Medicare
+supplemental coverage.
+- **Optum De-Identified Clinformatics® Data Mart Database – Date of Death
+  (Optum-DOD)** is a claims database including members of private health
+insurance.
+- **Optum de-identified Electronic Health Record Dataset (Optum-EHR)** is an
+  EHR database including Humedica’s Electronic Health Record.
 
-- **Optum de-identified Electronic Health Record Dataset (Optum-EHR)** is an EHR database including Humedica’s Electronic Health Record.
+Our cohorts included female new users of teriparatide or oral bisphosphonates
+above the age of 65 [@Ryan2013]. Patients were considered to be new-users if
+they received no anti-osteoporosis drugs (raloxifene, bazedoxifene, denosumab,
+abaloparatide, romosozumab) 365 days prior to treatment initiation with
+teriparatide or an oral bisphosphonate. We required patients to have at least
+one year of observation period before first exposure and an established
+osteoporosis diagnosis, defined as any recorded fracture of the hip, wrist,
+spine or shoulder/humerus in their recorded medical history.
 
-Our cohorts included female new users of teriparatide or oral bisphosphonates above the age of 65 [@Ryan2013].
-Patients were considered to be new-users if they received no anti-osteoporosis drugs (raloxifene, bazedoxifene, denosumab, abaloparatide, romosozumab) 365 days prior to treatment initiation with teriparatide or an oral bisphosphonate.
-We required patients to have at least one year of observation period before first exposure and an established osteoporosis diagnosis, defined as any recorded fracture of the hip, wrist, spine or shoulder/humerus in their recorded medical history.
-
-Our primary efficacy outcome was hip fracture.
-Vertebral fracture and a composite major osteoporotic fracture, defined as hip, vertebral or wrist/forearm/proximal humerus fracture, were our secondary efficacy outcomes.
-We excluded patients that had experienced the outcome under study any time prior to treatment initiation.
-Patient time-at-risk started 1 day after treatment initiation and finished 730 days after treatment initiation.
+Our primary efficacy outcome was hip fracture. Vertebral fracture and a
+composite major osteoporotic fracture, defined as hip, vertebral or
+wrist/forearm/proximal humerus fracture, were our secondary efficacy outcomes.
+We excluded patients that had experienced the outcome under study any time
+prior to treatment initiation. Patient time-at-risk started 1 day after
+treatment initiation and finished 730 days after treatment initiation.
 
 ##     Statistical analyses
-We carried out two sets of analyses.
-First, we derived overall treatment effect estimates of teriparatide compared to oral bisphosphonates regarding the three outcomes of interest.
-To account for potential measured confounding we developed separate large-scale propensity score models within each database based on LASSO logistic regression using the same predefined set of measured covariates [@Tian2018].
-We estimated treatment effects using Cox proportional hazards models with treatment as the sole covariate fitted in the 1:4 propensity score-matched subset of the considered patient population.
-Results from different databases were summarized using random effects meta-analysis.
+We carried out two sets of analyses. First, we derived overall treatment effect
+estimates of teriparatide compared to oral bisphosphonates regarding the three
+outcomes of interest. To account for potential measured confounding we
+developed separate large-scale propensity score models within each database
+based on LASSO logistic regression using the same predefined set of measured
+covariates [@Tian2018]. We estimated treatment effects using Cox proportional
+hazards models with treatment as the sole covariate fitted in the 1:4
+propensity score-matched subset of the considered patient population. Results
+from different databases were summarized using random effects meta-analysis.
 
-For the second set of analyses, we used the standardized framework for risk-based assessment of treatment effect heterogeneity, that follows the directives of the PATH statement [@Kent2019; @KentEnE2019].
-Following data extraction, the framework was applied in two steps.
-First, we derived individualized risk predictions for the three efficacy outcomes.
-We built the prediction models using LASSO logistic regression on the propensity score matched (1:4) subpopulation of the pooled treatment arms, aiming to remove any effect of measured confounding that may cause differential fit of our models across treatment arms.
-We considered the same large set of candidate covariates as for the development of the propensity score models.
+For the second set of analyses, we used the standardized framework for
+risk-based assessment of treatment effect heterogeneity, that follows the
+directives of the PATH statement [@Kent2019; @KentEnE2019]. Following data
+extraction, the framework was applied in two steps. First, we derived
+individualized risk predictions for the three efficacy outcomes. We built the
+prediction models using LASSO logistic regression on the propensity score
+matched (1:4) subpopulation of the pooled treatment arms, aiming to remove any
+effect of measured confounding that may cause differential fit of our models
+across treatment arms. We considered the same large set of candidate covariates
+as for the development of the propensity score models.
 
-For each outcome we used the derived prediction models to divide the population in 4 equally-sized subgroups, based on the quarters of the predicted risk distribution.
-Within each of these risk-based subgroups we developed a new propensity score model.
-Our analyses were performed on the propensity score matched (1:4) subset of the risk subgroup subset.
-We derived relative effect estimates using Cox proportional hazards models only with treatment as a predictor.
-Absolute effect estimates were calculated based on the difference of the Kaplan-Meier estimates, on day 730 after treatment initiation.
+For each outcome we used the derived prediction models to divide the population
+in 4 equally-sized subgroups, based on the quarters of the predicted risk
+distribution. Within each of these risk-based subgroups we developed a new
+propensity score model. Our analyses were performed on the propensity score
+matched (1:4) subset of the risk subgroup subset. We derived relative effect
+estimates using Cox proportional hazards models only with treatment as a
+predictor. Absolute effect estimates were calculated based on the difference of
+the Kaplan-Meier estimates, on day 730 after treatment initiation.
 
-Residual study bias from unmeasured confounding can still be present in observational studies, which often may not be visible when evaluating propensity score adjustment.
-To account for that, we considered a set of 126 negative control outcome experiments, where a null effect was assumed to be true.
-We used the estimated relative effects for these outcomes to derive an empirical approximation to the true null distribution, which was then used to calibrate the hazard ratios and their 95% confidence intervals for the 3 outcomes of interest [@Schuemie2013; @Schuemie2016].
+Residual study bias from unmeasured confounding can still be present in
+observational studies, which often may not be visible when evaluating
+propensity score adjustment. To account for that, we considered a set of 126
+negative control outcome experiments, where a null effect was assumed to be
+true. We used the estimated relative effects for these outcomes to derive an
+empirical approximation to the true null distribution, which was then used to
+calibrate the hazard ratios and their 95% confidence intervals for the 3
+outcomes of interest [@Schuemie2013; @Schuemie2016].
 
@@
 <!-- Need to fix the numbers. See icpe abstract -->
-A total of `r sum(incidenceOverall$maxTreatment) + sum(incidenceOverall$maxComparator)` patients were initially considered for our analyses.
-Of those, `r sum(incidenceOverall$maxTreatment)` received teriparatide and `r sum(incidenceOverall$maxComparator)` received an oral bisphosphonate \@ref(tab:table1).
+A total of `r sum(incidenceOverall$maxTreatment) +
+sum(incidenceOverall$maxComparator)` patients were initially considered for our
+analyses. Of those, `r sum(incidenceOverall$maxTreatment)` received
+teriparatide and `r sum(incidenceOverall$maxComparator)` received an oral
+bisphosphonate (Table\@ref(tab:table1)).
 
@@
 
-The random effects meta-analysis of teriparatide compared to oral bisphosphonates across the 3 considered databases gave a calibrated overall hazard ratio of `r metaCalibrateOverall %>% filter(outcome == 101) %>% pull(hr)` (`r metaCalibrateOverall %>% filter(outcome == 101) %>% pull(lower)` to `r metaCalibrateOverall %>% filter(outcome == 101) %>% pull(upper)`; 95% CI) for hip fracture, `r metaCalibrateOverall %>% filter(outcome == 102) %>% pull(hr)` (`r metaCalibrateOverall %>% filter(outcome == 102) %>% pull(lower)` to `r metaCalibrateOverall %>% filter(outcome == 102) %>% pull(upper)`; 95%CI) for major osteoporotic fracture and `r metaCalibrateOverall %>% filter(outcome == 103) %>% pull(hr)` (`r metaCalibrateOverall %>% filter(outcome == 103) %>% pull(lower)` to `r metaCalibrateOverall %>% filter(outcome == 103) %>% pull(upper)`; 95%CI) for vertebral fracture (Figure \@ref(fig:plotMeta)).
+The random effects meta-analysis of teriparatide compared to oral
+bisphosphonates across the 3 considered databases gave a calibrated overall
+hazard ratio of `r metaCalibrateOverall %>% filter(outcome == 101) %>%
+pull(hr)` (`r metaCalibrateOverall %>% filter(outcome == 101) %>% pull(lower)`
+to `r metaCalibrateOverall %>% filter(outcome == 101) %>% pull(upper)`; 95% CI)
+for hip fracture, `r metaCalibrateOverall %>% filter(outcome == 102) %>%
+pull(hr)` (`r metaCalibrateOverall %>% filter(outcome == 102) %>% pull(lower)`
+to `r metaCalibrateOverall %>% filter(outcome == 102) %>% pull(upper)`; 95%CI)
+for major osteoporotic fracture and `r metaCalibrateOverall %>% filter(outcome
+== 103) %>% pull(hr)` (`r metaCalibrateOverall %>% filter(outcome == 103) %>%
+pull(lower)` to `r metaCalibrateOverall %>% filter(outcome == 103) %>%
+pull(upper)`; 95%CI) for vertebral fracture (Figure \@ref(fig:plotMeta)).
 

### 2021-10-14 submission/manuscript.rmd
rmemove all text and restart methods

@@
 ## Study design
+
+We performed a retrospective new user cohort study to estimate the effectiveness
+of teriparatide compared to oral bisphosphonates in patients with
+osteoporosis [@Ryan2013]. We included female participants above the age of 50
+with established osteoporosis (any condition occurrence of hip, wrist, spine or
+shoulder/humerus fracture in their history, prior to treatment initiation) and
+at least 365 days of continuous observation period before the index event.
+Patients were considered to be new-users if they received no anti-osteoporosis
+drugs (raloxifene, bazedoxifene, denosumab, abaloparatide, romosozumab) 365 days
+prior to treatment initiation with teriparatide or an oral bisphosphonate.
+
+As a secondary analysis, we evaluated treatment effect heterogeneity using a risk-based approach. The
+
+Our primary efficacy outcome was hip fracture. Vertebral fracture and a
+composite major osteoporotic fracture, defined as hip, vertebral or
+wrist/forearm/proximal humerus fracture, were our secondary efficacy outcomes.
+We excluded patients that had experienced the outcome under study any time
+prior to treatment initiation. Patient time-at-risk started 1 day after
+treatment initiation and finished 730 days after treatment initiation.
+
+## Data sources
+
 We ran our analyses on three US observational databases mapped to OMOP-CDM
@@
   EHR database including Humedica’s Electronic Health Record.
-
-Our cohorts included female new users of teriparatide or oral bisphosphonates
-above the age of 65 [@Ryan2013]. Patients were considered to be new-users if
-they received no anti-osteoporosis drugs (raloxifene, bazedoxifene, denosumab,
-abaloparatide, romosozumab) 365 days prior to treatment initiation with
-teriparatide or an oral bisphosphonate. We required patients to have at least
-one year of observation period before first exposure and an established
-osteoporosis diagnosis, defined as any recorded fracture of the hip, wrist,
-spine or shoulder/humerus in their recorded medical history.
-
-Our primary efficacy outcome was hip fracture. Vertebral fracture and a
-composite major osteoporotic fracture, defined as hip, vertebral or
-wrist/forearm/proximal humerus fracture, were our secondary efficacy outcomes.
-We excluded patients that had experienced the outcome under study any time
-prior to treatment initiation. Patient time-at-risk started 1 day after
-treatment initiation and finished 730 days after treatment initiation.
 
@@
 
-```{r, echo=FALSE}
-incidenceOverall <- readRDS(here::here("data/raw/incidenceOverall.rds")) %>%
-  dplyr::filter(analysisType == "matchOnPs_1_to_4") %>%
-  dplyr::group_by(database) %>%
-  dplyr::summarise(
-    maxTreatment = max(treatmentPersons),
-    maxComparator = max(comparatorPersons),
-    .groups = "drop"
-  )
-```
-
-<!-- Need to fix the numbers. See icpe abstract -->
-A total of `r sum(incidenceOverall$maxTreatment) +
-sum(incidenceOverall$maxComparator)` patients were initially considered for our
-analyses. Of those, `r sum(incidenceOverall$maxTreatment)` received
-teriparatide and `r sum(incidenceOverall$maxComparator)` received an oral
-bisphosphonate (Table\@ref(tab:table1)).
-
-```{r, echo=FALSE}
-metaCalibrateOverall <- readRDS(
-  here::here("data/processed/metaCalibrateOverall.rds")
-) %>%
-  mutate(
-    hr = d(hr),
-    lower = d(lower),
-    upper = d(upper)
-  )
-```
-
-The random effects meta-analysis of teriparatide compared to oral
-bisphosphonates across the 3 considered databases gave a calibrated overall
-hazard ratio of `r metaCalibrateOverall %>% filter(outcome == 101) %>%
-pull(hr)` (`r metaCalibrateOverall %>% filter(outcome == 101) %>% pull(lower)`
-to `r metaCalibrateOverall %>% filter(outcome == 101) %>% pull(upper)`; 95% CI)
-for hip fracture, `r metaCalibrateOverall %>% filter(outcome == 102) %>%
-pull(hr)` (`r metaCalibrateOverall %>% filter(outcome == 102) %>% pull(lower)`
-to `r metaCalibrateOverall %>% filter(outcome == 102) %>% pull(upper)`; 95%CI)
-for major osteoporotic fracture and `r metaCalibrateOverall %>% filter(outcome
-== 103) %>% pull(hr)` (`r metaCalibrateOverall %>% filter(outcome == 103) %>%
-pull(lower)` to `r metaCalibrateOverall %>% filter(outcome == 103) %>%
-pull(upper)`; 95%CI) for vertebral fracture (Figure \@ref(fig:plotMeta)).
-
-```{r, echo=FALSE} 
-metaCalibrateRiskStratified <- readRDS(
-  here::here("data/processed/metaCalibrateRiskStratified.rds")
-) %>%
-  mutate(
-    hr = d(hr),
-    lower = d(lower),
-    upper = d(upper)
-  )
-
-calibrateRiskStratified <- readRDS(
-  here::here("data/processed/calibrateRiskStratified.rds")
-) %>%
-  mutate(
-    hr = d(exp(logRr)),
-    lower = d(exp(logLb95Rr)),
-    upper = d(exp(logUb95Rr))
-  )
-
-hipFractureAbsolute <- readRDS(
-  here::here("data/processed/hipFractureAbsolute.rds")
-) %>%
-  filter(riskStratum == "Q4") %>%
-  mutate(
-    estimate = d(100 * estimate),
-    lower    = d(100 * lower),
-    upper    = d(100 * upper)
-  )
-
-hipRiskPerformance <- readRDS(here::here("data/raw/predictionPerformance.rds")) %>%
-  filter(
-    cohort == "EntirePopulation",
-    stratOutcome == 101
-  )
-
-```
-
-The internally developed prediction models for hip fracture showed adequate performance achieving AUC of `r hipRiskPerformance %>% filter(database == "mdcr") %>% pull(auc) %>% d()`, `r hipRiskPerformance %>% filter(database == "optum_dod") %>% pull(auc) %>% d()` and `r hipRiskPerformance %>% filter(database == "panther") %>% pull(auc) %>% d()` in MDCR, Optum-DOD and Optum-EHR respectively.
-When assessing heterogeneity of treatment effect for hip fracture, we estimated calibrated hazard ratios of `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "mdcr") %>% pull(hr)` (`r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "mdcr") %>% pull(lower)` to `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "mdcr") %>% pull(upper)`; 95% CI),
-`r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "optum_dod") %>% pull(hr)` (`r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "optum_dod") %>% pull(lower)` to `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "optum_dod") %>% pull(upper)`; 95% CI) and `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "panther") %>% pull(hr)` (`r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "panther") %>% pull(lower)` to `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "panther") %>% pull(upper)`; 95% CI) within the highest risk quarter of MDCR, Optum-DOD and Optum-EHR respectively.
-This translated in absolute risk reduction of `r hipFractureAbsolute %>% filter(database == "mdcr") %>% pull(estimate)`% (`r hipFractureAbsolute %>% filter(database == "mdcr") %>% pull(lower)`% to `r hipFractureAbsolute %>% filter(database == "mdcr") %>% pull(upper)`%; 95% CI), `r hipFractureAbsolute %>% filter(database == "optum_dod") %>% pull(estimate)`% (`r hipFractureAbsolute %>% filter(database == "optum_dod") %>% pull(lower)`% to `r hipFractureAbsolute %>% filter(database == "optum_dod") %>% pull(upper)`%; 95% CI), `r hipFractureAbsolute %>% filter(database == "panther") %>% pull(estimate)`% (`r hipFractureAbsolute %>% filter(database == "panther") %>% pull(lower)`% to `r hipFractureAbsolute %>% filter(database == "panther") %>% pull(upper)`%; 95% CI) respectively (Figure \@ref(fig:plotAbsoluteHip)).
-The random effects meta-analytic calibrated hazard ratio within the highest risk quarter was `r metaCalibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101) %>% pull(hr)` (`r metaCalibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101) %>% pull(lower)` to `r metaCalibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101) %>% pull(upper)`; 95% CI) across the 3 databases.
-For the secondary outcomes (major osteoporotic fracture and vertebral fracture) we found no evidence of risk-based treatment effect heterogeneity.
-This was anticipated as overall effects of teriparatide on theses outcomes were not significant at the 95% confidence level.
-The full set of results can be explored at https://arekkas.shinyapps.io/ter_bis_3dbs/.
-
-<!-- How much sense does it make to look at meta-analytic hazard ratios in risk-quarters? Different models were used in each case  -->
-
-<!-- For the assessment of treatment effect heterogeneity none of the internally developed prediction models achieved high discrimination, with the models predicting the composite major osteoporotic fracture suffering the most (##Table 2).
-However, models on hip fracture--the primary efficacy outcome--displayed adequate performance.
-More information on the performance of the prediction models can be found in the supplement.
--->
-
-<!-- When analyzing the absolute risk difference of teriparatide compared to oral bisphosphonates within strata of predicted hip fracture risk we did not observe any clear pattern except for a non-significant absolute benefit for hip fracture within the highest risk patients (Figure 4), namely 0.55% (-0.89 to 1.98; 95%CI) in MDCR, 0.97% (-0.27 to 2.22; 95%CI) in OPTUM_DOD and 0.45% (-0.56 to 1.46; 95%CI) in PANTHER.
-Absolute and relative treatment effect estimates within strata of predicted risk of the secondary outcomes are available in the supplement.
--->
-
 \newpage
@@
 # Tables and figures
-```{r table1, echo=FALSE, results='asis'}
-incidenceOverall <- readRDS(here::here("data/raw/incidenceOverall.rds"))
-incidence <- incidenceOverall %>% 
-  filter(analysisType == "matchOnPs_1_to_4") %>%
-  mutate(
-    database = case_when(
-      database == "mdcr" ~ "MDCR",
-      database == "optum_dod" ~ "Optum-DOD",
-      database == "panther" ~ "Optum-EHR"
-    ),
-    treatmentDays = round(treatmentDays / 365),
-    comparatorDays = round(comparatorDays / 365)
-  ) %>%
-  select(
-    database,
-    outcomeId,
-    treatmentPersons,
-    treatmentDays,
-    treatmentOutcomes,
-    comparatorPersons,
-    comparatorDays,
-    comparatorOutcomes
-  ) %>%
-  arrange(database, outcomeId) %>%
-  mutate(
-    database = ifelse(
-      outcomeId == 101,
-      database,
-      ""
-    ),
-    outcomeId = case_when(
-      outcomeId == 101 ~ "Hip fracture",
-      outcomeId == 102 ~ "Major osteoporotic\nfracture",
-      outcomeId == 103 ~ "Vertebral fracture"
-    ),
-  )
-
-knitr::kable(
-  incidence,
-  format   = "latex",
-  longtable = TRUE,
-  booktabs = TRUE,
-  col.names = c(
-    "Database",
-    "Outcome",
-    "Persons",
-    "Person years",
-    "Events",
-    "Persons",
-    "Person years",
-    "Events"
-  ),
-  format.args = list(
-    big.mar = ","
-  ),
-   caption = "Number of patients, person years and outcomes for each outcome across all databases."
-) %>%
-  add_header_above(c("", "", "Teriparatide" = 3, "Bisphosphonates" = 3))
-  # kableExtra::add_indent(c(1, 4, 7))
-```
-
-\newpage
-```{r plotMeta, echo=FALSE, fig.cap="Calibrated hazard ratios for the 3 outcomes of interest across the 3 considered databases using a set of 126 negative controls.
-Values below 1 favour teriparatide, while values above 1 favour bisphosphonates.", out.width = '100%'}
-knitr::include_graphics(here::here("figures/plotMeta.png"))
-```
-
-\newpage
-```{r plotAbsoluteHip, echo=FALSE, fig.cap="Absolute risk differences wihtin quarters of predicted hip fracture risk across the 3 considered databases.", out.width = '100%'}
-knitr::include_graphics(here::here("figures/plotAbsoluteHip.png"))
-```

### 2021-11-04 submission/manuscript.rmd
writing results

@@
 library(kableExtra)
+
+options(scipen=1, digits=2)
 
@@
 # Introduction
-Osteoporosis is a common condition characterised by decreased bone density and
-associated with increased risk for fragility fractures, which affects almost
-30% of women aged $\geq50$ years [@wright2017impact]. It was estimated that in
-2005, over 2 million incident fractures occurred in the United States, with hip
-fractures accounting for more than 70% of the costs. By 2025, annual fractures
+Osteoporosis is a common condition characterised by decreased bone
+density and associated with increased risk for fragility fractures,
+which affects almost 30% of women aged $\geq50$ years
+[@wright2017impact]. It was estimated that in 2005, over 2 million
+incident fractures occurred in the United States, with hip fractures
+accounting for more than 70% of the costs. By 2025, annual fractures
 and costs could rise by almost 50% [@Cole2008].
 
-Fracture prevention is thus the key focus of anti-osteoporotic therapy,
-reducing the disease burden both for the affected patient but also on the
-population level. Several pharmacological agents are available for primary and
-secondary prevention. The choice of anti-osteoporotic agent largely depends on
-history of fragility fracture and anticipated fracture risk. Oral
-bisphosphonates (BP) are first line treatments for postmenopausal patients with
-increased fracture risk in clinical guidelines (#AACE/ACE 2020) as for their
-favourable cost-effectiveness and safety profile. Teriparatide, a parathyroid
-hormone analogue administered as a daily injection, was approved by the FDA in
-2002 as the first anabolic agent for treatment of severe postmenopausal
-osteoporosis. While clinical trials among patients with previous vertebral
-fractures showed a substantial risk reduction for new vertebral fracture
-comparing teriparatide to placebo [@Neer2001] and risedronate [@Kendler2018],
-its efficacy on low-incident major osteoporotic fractures, especially hip
-fractures, is less well established: Previous randomised trials assessing hip
-fracture comprised only few events, thus not providing sufficient power to show
-differences between treatment groups when studying comparative effectiveness. A
-recent meta-analysis by Diez-Perez and colleagues assessed the effect of
-teriparatide on hip fracture, indicating a significant 80% risk reduction
-compared to placebo and a non-significant 46% risk reduction when compared to
-active controls [@DezPrez2019].
+Fracture prevention is thus the key focus of anti-osteoporotic
+therapy, reducing the disease burden both for the affected patient but
+also on the population level. Several pharmacological agents are
+available for primary and secondary prevention. The choice of
+anti-osteoporotic agent largely depends on history of fragility
+fracture and anticipated fracture risk. Oral bisphosphonates (BP) are
+first line treatments for postmenopausal patients with increased
+fracture risk in clinical guidelines (#AACE/ACE 2020) as for their
+favourable cost-effectiveness and safety profile. Teriparatide, a
+parathyroid hormone analogue administered as a daily injection, was
+approved by the FDA in 2002 as the first anabolic agent for treatment
+of severe postmenopausal osteoporosis. While clinical trials among
+patients with previous vertebral fractures showed a substantial risk
+reduction for new vertebral fracture comparing teriparatide to placebo
+[@Neer2001] and risedronate [@Kendler2018], its efficacy on
+low-incident major osteoporotic fractures, especially hip fractures,
+is less well established: Previous randomised trials assessing hip
+fracture comprised only few events, thus not providing sufficient
+power to show differences between treatment groups when studying
+comparative effectiveness. A recent meta-analysis by Diez-Perez and
+colleagues assessed the effect of teriparatide on hip fracture,
+indicating a significant 80% risk reduction compared to placebo and a
+non-significant 46% risk reduction when compared to active controls
+[@DezPrez2019].
 
-With teriparatide-containing biosimilars being launched in recent years,
-treatment costs dropped significantly. Subsequently, the discussion of
-cost-effectiveness in patients with less severe forms of osteoporosis is
-restarting. Therefore, evaluation of the effectiveness in hip fracture
-prevention is required. In particular, observational studies assessing
-comparative effectiveness in a real-world setting are required, as
-effectiveness may differ from clinical trial as for differences in patient’s
-persistence and adherence to their anti-osteoporotic treatment. While hip
-fracture is a comparatively rare outcome, it is unambiguously defined and
-reliably recorded in routinely collected data, allowing for a direct comparison
-to results from clinical trial meta-analyses. While results from
-non-controlled, observational studies suggest that teriparatide may reduce the
-risk of hip fractures [@Burge2016; @Silverman2018], this is the first
-observational study assessing teriparatide in hip fracture risk reduction
-compared to oral bisphosphonate users in postmenopausal women using multiple
-large real-world databases.
+With teriparatide-containing biosimilars being launched in recent
+years, treatment costs dropped significantly. Subsequently, the
+discussion of cost-effectiveness in patients with less severe forms of
+osteoporosis is restarting. Therefore, evaluation of the effectiveness
+in hip fracture prevention is required. In particular, observational
+studies assessing comparative effectiveness in a real-world setting
+are required, as effectiveness may differ from clinical trial as for
+differences in patient’s persistence and adherence to their
+anti-osteoporotic treatment. While hip fracture is a comparatively
+rare outcome, it is unambiguously defined and reliably recorded in
+routinely collected data, allowing for a direct comparison to results
+from clinical trial meta-analyses. While results from non-controlled,
+observational studies suggest that teriparatide may reduce the risk of
+hip fractures [@Burge2016; @Silverman2018], this is the first
+observational study assessing teriparatide in hip fracture risk
+reduction compared to oral bisphosphonate users in postmenopausal
+women using multiple large real-world databases.
 
@@
 
-We performed a retrospective new user cohort study to estimate the effectiveness
-of teriparatide compared to oral bisphosphonates in patients with
-osteoporosis [@Ryan2013]. We included female participants above the age of 50
-with established osteoporosis (any condition occurrence of hip, wrist, spine or
-shoulder/humerus fracture in their history, prior to treatment initiation) and
-at least 365 days of continuous observation period before the index event.
-Patients were considered to be new-users if they received no anti-osteoporosis
-drugs (raloxifene, bazedoxifene, denosumab, abaloparatide, romosozumab) 365 days
-prior to treatment initiation with teriparatide or an oral bisphosphonate.
+We performed a retrospective new user cohort study to estimate the
+effectiveness of teriparatide compared to oral bisphosphonates in
+patients with osteoporosis [@Ryan2013]. We included female
+participants above the age of 50 with established osteoporosis (any
+condition occurrence of hip, wrist, spine or shoulder/humerus fracture
+in their history, prior to treatment initiation) and at least 365 days
+of continuous observation period before the index event.  Patients
+were considered to be new-users if they received no anti-osteoporosis
+drugs (raloxifene, bazedoxifene, denosumab, abaloparatide,
+romosozumab) 365 days prior to treatment initiation with teriparatide
+or an oral bisphosphonate.
 
-As a secondary analysis, we evaluated treatment effect heterogeneity using a risk-based approach. The
+We evaluated treatment effect heterogeneity using two different
+risk-based subgrouping approaches. After internally developing a hip
+fracture risk model within each database, we first assessed relative
+and absolute treatment effects in patients falling in the lowest 75%
+risk and the highest 25% risk subgroups. As a secondary approach, we
+separated patients based on current guidelines for targeting
+teriparatide treatment.... (EXPAND!!!!)
 
-Our primary efficacy outcome was hip fracture. Vertebral fracture and a
-composite major osteoporotic fracture, defined as hip, vertebral or
-wrist/forearm/proximal humerus fracture, were our secondary efficacy outcomes.
-We excluded patients that had experienced the outcome under study any time
-prior to treatment initiation. Patient time-at-risk started 1 day after
-treatment initiation and finished 730 days after treatment initiation.
+Our primary efficacy outcome was hip fracture. Vertebral fracture and
+a composite major osteoporotic fracture, defined as hip, vertebral or
+wrist/forearm/proximal humerus fracture, were our secondary efficacy
+outcomes.  We excluded patients that had experienced the outcome under
+study any time prior to treatment initiation. Patient time-at-risk
+started 1 day after treatment initiation and finished 730 days after
+treatment initiation.
 
@@
 
-- **IBM MarketScan Medicare Supplemental Database (MDCR)** is a claims database
+- **IBM MarketScan Medicare Supplemental Database (CCAE)** is a claims database
   representing retirees in the United States with primary or Medicare
@@
 ##     Statistical analyses
-We carried out two sets of analyses. First, we derived overall treatment effect
-estimates of teriparatide compared to oral bisphosphonates regarding the three
-outcomes of interest. To account for potential measured confounding we
-developed separate large-scale propensity score models within each database
-based on LASSO logistic regression using the same predefined set of measured
-covariates [@Tian2018]. We estimated treatment effects using Cox proportional
+We carried out two sets of analyses. First, we derived overall
+treatment effect estimates of teriparatide compared to oral
+bisphosphonates regarding the three outcomes of interest. To account
+for potential measured confounding we developed separate large-scale
+propensity score models within each database based on LASSO logistic
+regression using the same predefined set of measured covariates
+[@Tian2018]. We estimated treatment effects using Cox proportional
 hazards models with treatment as the sole covariate fitted in the 1:4
-propensity score-matched subset of the considered patient population. Results
-from different databases were summarized using random effects meta-analysis.
+propensity score-matched subset of the considered patient
+population. Hazard ratios (HR) derived in different databases were
+summarized using random effects meta-analysis.
 
 For the second set of analyses, we used the standardized framework for
-risk-based assessment of treatment effect heterogeneity, that follows the
-directives of the PATH statement [@Kent2019; @KentEnE2019]. Following data
-extraction, the framework was applied in two steps. First, we derived
-individualized risk predictions for the three efficacy outcomes. We built the
-prediction models using LASSO logistic regression on the propensity score
-matched (1:4) subpopulation of the pooled treatment arms, aiming to remove any
-effect of measured confounding that may cause differential fit of our models
-across treatment arms. We considered the same large set of candidate covariates
-as for the development of the propensity score models.
+risk-based assessment of treatment effect heterogeneity, that follows
+the directives of the PATH statement [@Kent2019; @KentEnE2019].
+Following data extraction, the framework was applied in
+two steps. First, we derived individualized risk predictions for the
+three efficacy outcomes. We built the prediction models using LASSO
+logistic regression on the propensity score matched (1:4)
+subpopulation of the pooled treatment arms, aiming to remove any
+effect of measured confounding that may cause differential fit of our
+models across treatment arms. We considered the same large set of
+candidate covariates as for the development of the propensity score
+models.
 
-For each outcome we used the derived prediction models to divide the population
-in 4 equally-sized subgroups, based on the quarters of the predicted risk
-distribution. Within each of these risk-based subgroups we developed a new
-propensity score model. Our analyses were performed on the propensity score
-matched (1:4) subset of the risk subgroup subset. We derived relative effect
-estimates using Cox proportional hazards models only with treatment as a
-predictor. Absolute effect estimates were calculated based on the difference of
-the Kaplan-Meier estimates, on day 730 after treatment initiation.
+For each outcome we used the derived prediction models to divide the
+population in 4 equally-sized subgroups, based on the quarters of the
+predicted risk distribution. Within each of these risk-based subgroups
+we developed a new propensity score model. Our analyses were performed
+on the propensity score matched (1:4) subset of the risk subgroup
+subset. We derived relative effect estimates using Cox proportional
+hazards models only with treatment as a predictor. Absolute effect
+estimates were calculated based on the difference of the Kaplan-Meier
+estimates, on day 730 after treatment initiation.
 
-Residual study bias from unmeasured confounding can still be present in
-observational studies, which often may not be visible when evaluating
-propensity score adjustment. To account for that, we considered a set of 126
-negative control outcome experiments, where a null effect was assumed to be
-true. We used the estimated relative effects for these outcomes to derive an
-empirical approximation to the true null distribution, which was then used to
-calibrate the hazard ratios and their 95% confidence intervals for the 3
-outcomes of interest [@Schuemie2013; @Schuemie2016].
+Residual study bias from unmeasured confounding can still be present
+in observational studies, which often may not be visible when
+evaluating propensity score adjustment. To account for that, we
+considered a set of 126 negative control outcome experiments, where a
+null effect was assumed to be true. We used the estimated relative
+effects for these outcomes to derive an empirical approximation to the
+true null distribution, which was then used to calibrate the hazard
+ratios and their 95% confidence intervals for the 3 outcomes of
+interest [@Schuemie2013; @Schuemie2016].
 
@@
 # Results
+We extracted information on a total of 28,657 patients receiving
+teriparatide and 1.02 million patients receiving an oral
+bisphosphonate across the 3 databases (Table 1).
+
+```{r, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
+calibrateOverallResults <- readRDS(here::here("data/processed/metaCalibrateOverall.rds"))
+```
+
+## Overall results
+In our overall analyses we found no difference between teriparatide
+and oral bisphosphonates in terms of hip fracture risk
+(HR 
+`r calibrateOverallResults %>% filter(outcome == "101") %>% pull(hr)`;
+`r calibrateOverallResults %>% filter(outcome == "101") %>% pull(lower)`
+to 
+`r calibrateOverallResults %>% filter(outcome == "101") %>% pull(upper)`,
+95% CI). We estimated major osteoporotic fracture risk increase (HR 
+`r calibrateOverallResults %>% filter(outcome == "102") %>% pull(hr)`;
+`r calibrateOverallResults %>% filter(outcome == "102") %>% pull(lower)` 
+to 
+`r calibrateOverallResults %>% filter(outcome == "102") %>% pull(upper)`,
+95% CI) and vertebral fracture risk increase (HR 
+`r calibrateOverallResults %>% filter(outcome == "103") %>% pull(hr)`;
+`r calibrateOverallResults %>% filter(outcome == "103") %>% pull(lower)`
+to `r calibrateOverallResults %>% filter(outcome == "103") %>% pull(upper)`,
+95% CI) with teriparatide (Figure \@ref(fig:plotMeta)).
+
+## Treatment effect heterogeneity
+
+```{r, echo=FALSE} 
+metaCalibrateRiskStratified_25_75 <- readRDS(
+  here::here("data/processed/metaCalibrateRiskStratified_age_50_tr_1_q_25_75.rds")
+) %>%
+  mutate(
+    hr = d(hr),
+    lower = d(lower),
+    upper = d(upper)
+  )
+calibrateRiskStratified_25_75 <- readRDS(
+  here::here("data/processed/calibrateRiskStratified_age_50_tr_1_q_25_75.rds")
+) %>%
+  mutate(
+    hr = d(exp(logRr)),
+    lower = d(exp(logLb95Rr)),
+    upper = d(exp(logUb95Rr))
+  )
+hipFractureAbsolute_age_50_tr_1_q_25_75_101_101 <- readRDS(
+  here::here("data/processed/hipFractureAbsolute_age_50_tr_1_q_25_75_101_101.rds")
+) %>%
+  filter(riskStratum == "Q2") %>%
+  mutate(
+    estimate = d(100 * estimate),
+    lower    = d(100 * lower),
+    upper    = d(100 * upper)
+  )
+
+hipFractureAbsolute_age_50_tr_1_q_25_75_101_102 <- readRDS(
+  here::here("data/processed/hipFractureAbsolute_age_50_tr_1_q_25_75_101_102.rds")
+) %>%
+  mutate(
+    estimate = d(100 * estimate),
+    lower    = d(100 * lower),
+    upper    = d(100 * upper)
+  )
+
+hipFractureAbsolute_age_50_tr_1_q_25_75_101_103 <- readRDS(
+  here::here("data/processed/hipFractureAbsolute_age_50_tr_1_q_25_75_101_103.rds")
+) %>%
+  mutate(
+    estimate = d(100 * estimate),
+    lower    = d(100 * lower),
+    upper    = d(100 * upper)
+  )
+
+hipRiskPerformance <- readRDS(here::here("data/raw/predictionPerformance.rds")) %>%
+  filter(
+    cohort == "EntirePopulation",
+    stratOutcome == 101
+  )
+```
+
+The internally developed prediction models for hip fracture showed adequate performance achieving AUC of 
+`r hipRiskPerformance %>% filter(database == "ccae") %>% pull(auc) %>% d()`,
+`r hipRiskPerformance %>% filter(database == "optum_extended_dod") %>% pull(auc) %>% d()`
+and `r hipRiskPerformance %>% filter(database == "optum_ehr") %>% pull(auc) %>% d()`
+in CCAE, Optum-DOD and Optum-EHR respectively. When assessing
+heterogeneity of treatment effect for hip fracture using a 75-25% risk
+split, we estimated calibrated hazard ratios of
+`r calibrateRiskStratified_25_75 %>% filter(riskStratum == "Q2", stratOutcome == 101, estOutcome == 101, database == "ccae") %>% pull(hr)` 
+(`r calibrateRiskStratified_25_75 %>% filter(riskStratum == "Q2", stratOutcome == 101, estOutcome == 101, database == "ccae") %>% pull(lower)` 
+to 
+`r calibrateRiskStratified_25_75 %>% filter(riskStratum == "Q2", stratOutcome == 101, estOutcome == 101, database == "ccae") %>% pull(upper)`;
+95% CI),
+`r calibrateRiskStratified_25_75 %>% filter(riskStratum == "Q2", stratOutcome == 101, estOutcome == 101, database == "optum_extended_dod") %>% pull(hr)`
+(`r calibrateRiskStratified_25_75 %>% filter(riskStratum == "Q2", stratOutcome == 101, estOutcome == 101, database == "optum_extended_dod") %>% pull(lower)`
+to 
+`r calibrateRiskStratified_25_75 %>% filter(riskStratum == "Q2", stratOutcome == 101, estOutcome == 101, database == "optum_extended_dod") %>% pull(upper)`;
+95% CI) and 
+`r calibrateRiskStratified_25_75 %>% filter(riskStratum == "Q2", stratOutcome == 101, estOutcome == 101, database == "optum_ehr") %>% pull(hr)`
+(`r calibrateRiskStratified_25_75 %>% filter(riskStratum == "Q2", stratOutcome == 101, estOutcome == 101, database == "optum_ehr") %>% pull(lower)`
+to
+`r calibrateRiskStratified_25_75 %>% filter(riskStratum == "Q2", stratOutcome == 101, estOutcome == 101, database == "optum_ehr") %>% pull(upper)`;
+95% CI) within the highest risk quarter of CCAE, Optum-DOD and Optum-EHR respectively. This translated in absolute risk reduction of
+`r hipFractureAbsolute_age_50_tr_1_q_25_75_101_101 %>% filter(database == "ccae") %>% pull(estimate)`% (`r hipFractureAbsolute_age_50_tr_1_q_25_75_101_101 %>% filter(database == "ccae") %>% pull(lower)`%
+to
+`r hipFractureAbsolute_age_50_tr_1_q_25_75_101_101 %>% filter(database == "ccae") %>% pull(upper)`%;
+95% CI),
+`r hipFractureAbsolute_age_50_tr_1_q_25_75_101_101 %>% filter(database == "optum_extended_dod") %>% pull(estimate)`%
+(`r hipFractureAbsolute_age_50_tr_1_q_25_75_101_101 %>% filter(database == "optum_extended_dod") %>% pull(lower)`%
+to
+`r hipFractureAbsolute_age_50_tr_1_q_25_75_101_101 %>% filter(database == "optum_extended_dod") %>% pull(upper)`%;
+95% CI), 
+`r hipFractureAbsolute_age_50_tr_1_q_25_75_101_101 %>% filter(database == "optum_ehr") %>% pull(estimate)`%
+(`r hipFractureAbsolute_age_50_tr_1_q_25_75_101_101 %>% filter(database == "optum_ehr") %>% pull(lower)`%
+to `r hipFractureAbsolute_age_50_tr_1_q_25_75_101_101 %>% filter(database == "optum_ehr") %>% pull(upper)`%;
+95% CI) respectively (Figure \@ref(fig:plotAbsoluteHip)). The random
+effects meta-analytic calibrated hazard ratio within the highest risk
+quarter was
+`r metaCalibrateRiskStratified_25_75 %>% filter(riskStratum == "Q2", stratOutcome == 101, estOutcome == 101) %>% pull(hr)`
+(`r metaCalibrateRiskStratified_25_75 %>% filter(riskStratum == "Q2", stratOutcome == 101, estOutcome == 101) %>% pull(lower)`
+to 
+`r metaCalibrateRiskStratified_25_75 %>% filter(riskStratum == "Q2", stratOutcome == 101, estOutcome == 101) %>% pull(upper)`;
+95% CI) across the 3 databases.
+
+For major osteoporotic fractures we estimated 
+`r hipFractureAbsolute_age_50_tr_1_q_25_75_101_102 %>% filter(database == "ccae", riskStratum == "Q1") %>% pull(estimate)`% (`r hipFractureAbsolute_age_50_tr_1_q_25_75_101_102 %>% filter(database == "ccae", riskStratum == "Q1") %>% pull(lower)`%
+to
+`r hipFractureAbsolute_age_50_tr_1_q_25_75_101_102 %>% filter(database == "ccae", riskStratum == "Q1") %>% pull(upper)`%;
+95% CI),
+`r hipFractureAbsolute_age_50_tr_1_q_25_75_101_102 %>% filter(database == "optum_extended_dod", riskStratum == "Q1") %>% pull(estimate)`%
+(`r hipFractureAbsolute_age_50_tr_1_q_25_75_101_102 %>% filter(database == "optum_extended_dod", riskStratum == "Q1") %>% pull(lower)`%
+to
+`r hipFractureAbsolute_age_50_tr_1_q_25_75_101_102 %>% filter(database == "optum_extended_dod", riskStratum == "Q1") %>% pull(upper)`%;
+95% CI), 
+`r hipFractureAbsolute_age_50_tr_1_q_25_75_101_102 %>% filter(database == "optum_ehr", riskStratum == "Q1") %>% pull(estimate)`%
+(`r hipFractureAbsolute_age_50_tr_1_q_25_75_101_102 %>% filter(database == "optum_ehr", riskStratum == "Q1") %>% pull(lower)`%
+to `r hipFractureAbsolute_age_50_tr_1_q_25_75_101_102 %>% filter(database == "optum_ehr", riskStratum == "Q1") %>% pull(upper)`%;
+95% CI) in the lower 75% hip fracture risk patients. This was reduced to 
+`r hipFractureAbsolute_age_50_tr_1_q_25_75_101_102 %>% filter(database == "ccae", riskStratum == "Q2") %>% pull(estimate)`% (`r hipFractureAbsolute_age_50_tr_1_q_25_75_101_102 %>% filter(database == "ccae", riskStratum == "Q1") %>% pull(lower)`%
+to
+`r hipFractureAbsolute_age_50_tr_1_q_25_75_101_102 %>% filter(database == "ccae", riskStratum == "Q2") %>% pull(upper)`%;
+95% CI),
+`r hipFractureAbsolute_age_50_tr_1_q_25_75_101_102 %>% filter(database == "optum_extended_dod", riskStratum == "Q2") %>% pull(estimate)`%
+(`r hipFractureAbsolute_age_50_tr_1_q_25_75_101_102 %>% filter(database == "optum_extended_dod", riskStratum == "Q2") %>% pull(lower)`%
+to
+`r hipFractureAbsolute_age_50_tr_1_q_25_75_101_102 %>% filter(database == "optum_extended_dod", riskStratum == "Q2") %>% pull(upper)`%;
+95% CI), 
+`r hipFractureAbsolute_age_50_tr_1_q_25_75_101_102 %>% filter(database == "optum_ehr", riskStratum == "Q2") %>% pull(estimate)`%
+(`r hipFractureAbsolute_age_50_tr_1_q_25_75_101_102 %>% filter(database == "optum_ehr", riskStratum == "Q2") %>% pull(lower)`%
+to `r hipFractureAbsolute_age_50_tr_1_q_25_75_101_102 %>% filter(database == "optum_ehr", riskStratum == "Q2") %>% pull(upper)`%;
+95% CI),
+
+For the secondary outcomes (major
+osteoporotic fracture and vertebral fracture) we found no evidence of
+risk-based treatment effect heterogeneity. This was anticipated as
+overall effects of teriparatide on theses outcomes were not
+significant at the 95% confidence level. The full set of results can
+be explored at https://arekkas.shinyapps.io/ter_bis_3dbs/.
 
@@
 # Tables and figures
+
+\newpage
+```{r plotMeta, echo=FALSE, fig.cap="Calibrated hazard ratios for the 3 outcomes of interest across the 3 considered databases using a set of 126 negative controls.
+Values below 1 favour teriparatide, while values above 1 favour bisphosphonates.", out.width = '100%'}
+grid::grid.raster(tiff::readTIFF(here::here("figures/plotMeta.tiff")))
+```

### 2021-11-04 code/extractAbsoluteHip.R

@@ -8,16 +8,36 @@
 # Output:
 #     data/processed/hipFractureAbsolute.rds
     
-    
+args <- commandArgs(trailingOnly = TRUE)    
+args_stratOutcome <- args[1]
+args_estOutcome <- args[2]
+args_analysisType <- args[3]
+
 library(tidyverse)
 
 absoluteResults <- readRDS("data/raw/mappedOverallAbsoluteResults.rds") %>%
     tibble()
 
+fileName <- paste0(
+  paste(
+    "hipFractureAbsolute",
+    args_analysisType,
+    args_stratOutcome,
+    args_estOutcome,
+    sep = "_"
+  ),
+  ".rds"
+)
+
 absoluteResults %>%
     filter(
-        analysisType == "matchOnPs_1_to_4",
-        stratOutcome == 101,
-        estOutcome == 101
+        analysisType == args_analysisType,
+        stratOutcome == args_stratOutcome,
+        estOutcome == args_estOutcome
     ) %>%
-    saveRDS("data/processed/hipFractureAbsolute.rds")
+    saveRDS(
+      file.path(
+        "data/processed",
+        fileName
+      )
+    )

### 2021-12-01 submission/manuscript.rmd
make negative control plot

--- 2021-10-14 submission/manuscript.rmd
@@
 ## Study design
+We performed a retrospective new user cohort study to estimate the
+effectiveness of teriparatide compared to oral bisphosphonates in
+patients with osteoporosis[8]. We included female participants above
+the age of 50 with established osteoporosis (any condition occurrence
+of hip, wrist, spine or shoulder/humerus fracture in their history,
+prior to treatment initiation) and at least 365 days of continuous
+observation period before the index event. Patients were considered to
+be new-users if they received no anti-osteoporosis drugs (raloxifene,
+bazedoxifene, denosumab, abaloparatide, romosozumab) 365 days prior to
+treatment initiation with teriparatide or an oral bisphosphonate.
 
-We performed a retrospective new user cohort study to estimate the effectiveness
-of teriparatide compared to oral bisphosphonates in patients with
-osteoporosis [@Ryan2013]. We included female participants above the age of 50
-with established osteoporosis (any condition occurrence of hip, wrist, spine or
-shoulder/humerus fracture in their history, prior to treatment initiation) and
-at least 365 days of continuous observation period before the index event.
-Patients were considered to be new-users if they received no anti-osteoporosis
-drugs (raloxifene, bazedoxifene, denosumab, abaloparatide, romosozumab) 365 days
-prior to treatment initiation with teriparatide or an oral bisphosphonate.
-
-As a secondary analysis, we evaluated treatment effect heterogeneity using a risk-based approach. The
-
-Our primary efficacy outcome was hip fracture. Vertebral fracture and a
-composite major osteoporotic fracture, defined as hip, vertebral or
-wrist/forearm/proximal humerus fracture, were our secondary efficacy outcomes.
-We excluded patients that had experienced the outcome under study any time
-prior to treatment initiation. Patient time-at-risk started 1 day after
-treatment initiation and finished 730 days after treatment initiation.
+Our primary efficacy outcome was hip fracture. Vertebral fracture and
+a composite major osteoporotic fracture, defined as hip, vertebral or
+wrist/forearm/proximal humerus fracture, were our secondary efficacy
+outcomes. We excluded patients that had experienced the outcome under
+study any time prior to treatment initiation. Patient time-at-risk
+started 1 day after treatment initiation and finished 730 days after
+treatment initiation.
 
@@
 
-- **IBM MarketScan Medicare Supplemental Database (MDCR)** is a claims database
-  representing retirees in the United States with primary or Medicare
-supplemental coverage.
-- **Optum De-Identified Clinformatics® Data Mart Database – Date of Death
-  (Optum-DOD)** is a claims database including members of private health
-insurance.
-- **Optum de-identified Electronic Health Record Dataset (Optum-EHR)** is an
-  EHR database including Humedica’s Electronic Health Record.
+- **IMB MarketScan Commercial Database (CCAE)** includes health
+  insurance claims across the continuum of care (e.g. inpatient,
+  outpatient, outpatient pharmacy, carve-out behavioral healthcare) as
+  well as enrollment data from large employers and health plans across
+  the United States who provide private healthcare coverage for more
+  than 155 million employees, their spouses, and dependents. This
+  administrative claims database includes a variety of fee-
+  for-service, preferred provider organizations, and capitated health
+  plans.
+- **Optum De-Identified Clinformatics® Data Mart Database – Date of
+  Death (Optum-DOD)**: is derived from a database of administrative
+  health claims for members of large commercial and Medicare Advantage
+  health plans. The database includes approximately 17-19 million
+  annual covered lives, for a total of over 65 million unique lives
+  over a 12 year period (1/2007 through 12/2019). The population is
+  geographically diverse, spanning all 50 states.
+- **Optum de-identified Electronic Health Record Dataset (Optum-EHR)**
+  is derived from dozens of healthcare provider organizations in the
+  United States, that include more than 700 Hospitals and 7000
+  Clinics; treating more than 102 million patients receiving care in
+  the United States.
 
 ##     Statistical analyses
-We carried out two sets of analyses. First, we derived overall treatment effect
-estimates of teriparatide compared to oral bisphosphonates regarding the three
-outcomes of interest. To account for potential measured confounding we
-developed separate large-scale propensity score models within each database
-based on LASSO logistic regression using the same predefined set of measured
-covariates [@Tian2018]. We estimated treatment effects using Cox proportional
+We carried out two sets of analyses. First, we derived overall
+treatment effect estimates of teriparatide compared to oral
+bisphosphonates regarding the three outcomes of interest. To account
+for potential measured confounding we developed separate large-scale
+propensity score models within each database based on LASSO logistic
+regression using the same predefined set of measured covariates
+[@Tian2018]. We estimated treatment effects using Cox proportional
 hazards models with treatment as the sole covariate fitted in the 1:4
-propensity score-matched subset of the considered patient population. Results
-from different databases were summarized using random effects meta-analysis.
+propensity score-matched subset of the considered patient
+population. Hazard ratios (HR) derived in different databases were
+summarized using random effects meta-analysis.
 
 For the second set of analyses, we used the standardized framework for
-risk-based assessment of treatment effect heterogeneity, that follows the
-directives of the PATH statement [@Kent2019; @KentEnE2019]. Following data
-extraction, the framework was applied in two steps. First, we derived
-individualized risk predictions for the three efficacy outcomes. We built the
-prediction models using LASSO logistic regression on the propensity score
-matched (1:4) subpopulation of the pooled treatment arms, aiming to remove any
-effect of measured confounding that may cause differential fit of our models
-across treatment arms. We considered the same large set of candidate covariates
-as for the development of the propensity score models.
+risk-based assessment of treatment effect heterogeneity, that follows
+the directives of the PATH statement [@Kent2019; @KentEnE2019].
+Following data extraction, the framework was applied in
+two steps. First, we derived individualized risk predictions for the
+three efficacy outcomes. We built the prediction models using LASSO
+logistic regression on the propensity score matched (1:4)
+subpopulation of the pooled treatment arms, aiming to remove any
+effect of measured confounding that may cause differential fit of our
+models across treatment arms. We considered the same large set of
+candidate covariates as for the development of the propensity score
+models.
 
-For each outcome we used the derived prediction models to divide the population
-in 4 equally-sized subgroups, based on the quarters of the predicted risk
-distribution. Within each of these risk-based subgroups we developed a new
-propensity score model. Our analyses were performed on the propensity score
-matched (1:4) subset of the risk subgroup subset. We derived relative effect
-estimates using Cox proportional hazards models only with treatment as a
-predictor. Absolute effect estimates were calculated based on the difference of
-the Kaplan-Meier estimates, on day 730 after treatment initiation.
+For each outcome we used the derived prediction models to divide the
+population in 4 equally-sized subgroups, based on the quarters of the
+predicted risk distribution. Within each of these risk-based subgroups
+we developed a new propensity score model. Our analyses were performed
+on the propensity score matched (1:4) subset of the risk subgroup
+subset. We derived relative effect estimates using Cox proportional
+hazards models only with treatment as a predictor. Absolute effect
+estimates were calculated based on the difference of the Kaplan-Meier
+estimates, on day 730 after treatment initiation.
 
-Residual study bias from unmeasured confounding can still be present in
-observational studies, which often may not be visible when evaluating
-propensity score adjustment. To account for that, we considered a set of 126
-negative control outcome experiments, where a null effect was assumed to be
-true. We used the estimated relative effects for these outcomes to derive an
-empirical approximation to the true null distribution, which was then used to
-calibrate the hazard ratios and their 95% confidence intervals for the 3
-outcomes of interest [@Schuemie2013; @Schuemie2016].
+For each outcome we used the derived prediction models to divide the
+population in 4 equally-sized subgroups, based on the quarters of the
+predicted risk distribution. Within each of these risk-based subgroups
+we developed a new propensity score model. Our analyses were performed
+on the propensity score matched (1:4) subset of the risk subgroup
+subset. We derived relative effect estimates using Cox proportional
+hazards models only with treatment as a predictor. Absolute effect
+estimates were calculated based on the difference of the Kaplan-Meier
+estimates, on day 730 after treatment initiation.
+
+Residual study bias from unmeasured confounding can still be present
+in observational studies, which often may not be visible when
+evaluating propensity score adjustment. To account for that, we
+considered a set of 126 negative control outcome experiments, where a
+null effect was assumed to be true. We used the estimated relative
+effects for these outcomes to derive an empirical approximation to the
+true null distribution, which was then used to calibrate the hazard
+ratios and their 95% confidence intervals for the 3 outcomes of
+interest [@Schuemie2013; @Schuemie2016].
 
@@
 # Results
+## Treatment effect heterogeneity
 

### 2021-12-01 code/PlotNegativeControls.R

@@ -1,0 +1,49 @@
+#!/usr/bin/env Rscript
+
+# =================================
+# Description:
+# Input:
+# Output:
+# Depends:
+# =================================
+
+library(tidyverse)
+library(EmpiricalCalibration)
+
+positiveResults <- readRDS("data/raw/mappedOverallResults.rds") %>%
+  filter(
+    database == "ccae",
+    analysisType == "age_50_tr_1_m_1_10",
+    outcomeId == 101
+  ) %>%
+  mutate(logRr = log(estimate))
+
+negativeControls <- readRDS(
+  "data/raw/mappedOverallResultsNegativeControls.rds"
+) %>%
+  filter(
+    database == "ccae",
+    analysisType == "age_50_tr_1_m_1_10"
+  ) %>%
+  mutate(logRr = log(estimate))
+
+null <- fitNull(
+  logRr = negativeControls$logRr,
+  seLogRr = negativeControls$seLogRr)
+
+plot <- plotCalibrationEffect(
+  null = null,
+  logRrNegatives = negativeControls$logRr,
+  seLogRrNegatives = negativeControls$seLogRr,
+  logRrPositives = positiveResults$logRr,
+  seLogRrPositives = positiveResults$seLogRr
+)
+
+ggsave(
+  file.path("figures/overallNcPlot.tiff"),
+  plot = plot,
+  dpi = 1200,
+  width = 10,
+  height = 8,
+  compression = "lzw"
+)

### 2021-12-03 submission/manuscript.rmd
prepare results figures except risk stratified

@@
 library(kableExtra)
+library(here)
 
@@
 
+***[ Introductory stuff for Table 1 !! ]***
+
 \newpage
@@
 \newpage
+```{r overallCovariateBalance, cache = TRUE, echo = FALSE, fig.cap="Calibrated overall results"}
+grid::grid.raster(tiff::readTIFF(here("figures/OverallCovariateBalance.tiff")))
+```
+
+\newpage
+```{r overallPsDensity, cache = FALSE, echo = FALSE, fig.cap="Calibrated overall results"}
+grid::grid.raster(tiff::readTIFF(here("figures/OverallPsDensity.tiff")))
+```
+
+\newpage
+```{r overallNcPlot, cache = FALSE, echo = FALSE, fig.cap="Overall negative control plot."}
+grid::grid.raster(tiff::readTIFF(here("figures/OverallNcPlot.tiff")))
+```
+
+\newpage
 ```{r plotMeta, echo=FALSE, fig.cap="Calibrated hazard ratios for the 3 outcomes of interest across the 3 considered databases using a set of 126 negative controls.

### 2021-12-03 code/PlotCovariateBalance.R

@@ -1,0 +1,52 @@
+#!/usr/bin/env Rscript
+
+library(tidyverse)
+
+databases <- c("ccae", "optum_extended_dod", "optum_ehr")
+databaseLabels <- c("CCAE", "Optum-DOD", "Optum-EHR")
+pp <- list()
+
+for (i in seq_along(databases)) {
+  pattern <- paste0("^overall_bal.*", databases[i], ".*101*.rds")
+  pp[[i]] <- list.files(
+    path = "data/processed",
+    pattern = pattern,
+    full.names = TRUE
+  ) %>%
+    map(readRDS) %>%
+    bind_rows() %>%
+    tibble() %>%	
+    CohortMethod::plotCovariateBalanceScatterPlot(
+      beforeLabel = "",
+      afterLabel = databaseLabels[i]
+    ) +
+    theme_bw() +
+    theme(
+      plot.title   = element_blank(),
+      axis.title.x = element_blank(),
+      axis.text.x  = element_text(size = 22),
+      axis.text.y  = element_text(size = 22),
+      axis.title   = element_text(size = 30),
+      strip.text   = element_text(size = 25),
+      strip.background = element_rect(colour = "black", fill = NA)
+    )
+  
+  if (i != 1) {
+    pp[[i]] <- pp[[i]] +
+      theme(
+        strip.text       = element_blank(),
+        strip.background = element_blank()
+      )
+  }
+}
+
+plot <- gridExtra::grid.arrange(pp[[1]], pp[[2]], pp[[3]], nrow = 1)
+ggsave(
+  "figures/OverallCovariateBalance.tiff",
+  plot, 
+  compression = "lzw", 
+  width       = 700, 
+  height      = 350,
+  units       = "mm",
+  dpi         = 600
+)

### 2021-12-03 code/PlotNegativeControls.R

@@ -10,40 +10,62 @@
 library(tidyverse)
 library(EmpiricalCalibration)
 
-positiveResults <- readRDS("data/raw/mappedOverallResults.rds") %>%
-  filter(
-    database == "ccae",
-    analysisType == "age_50_tr_1_m_1_10",
-    outcomeId == 101
-  ) %>%
-  mutate(logRr = log(estimate))
+databases <- c("ccae", "optum_extended_dod", "optum_ehr")
+databaseLabels <- c("CCAE", "Optum-DOD", "Optum-EHR")
+mappedOverallResults <- readRDS("data/processed/mappedOverallResults.rds")
+overallNegativeControls <- readRDS("data/processed/mappedOverallResultsNegativeControls.rds")
+plot <- list()
 
-negativeControls <- readRDS(
-  "data/raw/mappedOverallResultsNegativeControls.rds"
-) %>%
-  filter(
-    database == "ccae",
-    analysisType == "age_50_tr_1_m_1_10"
-  ) %>%
-  mutate(logRr = log(estimate))
+for (i in seq_along(databases)) {
+  positiveResults <- mappedOverallResults %>%
+    filter(
+      database == databases[i],
+      analysisType == "age_50_tr_1_m_1_10",
+      outcomeId == 101
+    ) %>%
+    mutate(logRr = log(estimate))
+  
+  negativeControls <- overallNegativeControls %>%
+    filter(
+      database == databases[i],
+      analysisType == "age_50_tr_1_m_1_10"
+    ) %>%
+    mutate(logRr = log(estimate))
+  
+  null <- fitNull(
+    logRr = negativeControls$logRr,
+    seLogRr = negativeControls$seLogRr)
+  
+  plot[[i]] <- plotCalibrationEffect(
+    null = null,
+    logRrNegatives = negativeControls$logRr,
+    seLogRrNegatives = negativeControls$seLogRr,
+    logRrPositives = positiveResults$logRr,
+    seLogRrPositives = positiveResults$seLogRr
+  ) +
+    ggtitle(databaseLabels[i]) +
+    theme(
+      plot.title = element_text(size = 38),
+      axis.text.x = element_text(size = 22),
+      axis.title.x  = element_text(size = 30),
+    )
+  if (i == 1) {
+    plot[[i]] <- plot[[i]] +
+      theme(axis.title.y = element_text(size = 30))
+  } else {
[…]

### 2021-12-03 code/PlotPsDensity.R

@@ -1,0 +1,80 @@
+#!/usr/bin/env Rscript
+
+library(tidyverse)
+
+databases <- c("ccae", "optum_extended_dod", "optum_ehr")
+databaseLabels <- c("CCAE", "Optum-DOD", "Optum-EHR")
+pp <- list()
+
+for (i in seq_along(databases)) {
+  pattern <- paste0("^overall_psDensity.*", databases[i], ".*101*.rds")
+  pp[[i]] <- list.files(
+    path = "data/processed",
+    pattern = pattern,
+    full.names = TRUE
+  ) %>%
+    map(readRDS) %>%
+    bind_rows() %>%
+    mutate(
+      treatment = as.factor(treatment)
+    ) %>%
+    tibble() %>%	
+    ggplot2::ggplot(
+      ggplot2::aes(
+        x = x,
+        y = y
+      )
+    ) +
+    ggplot2::geom_density(
+      stat = "identity",
+      ggplot2::aes(
+        color = treatment,
+        group = treatment,
+        fill = treatment
+      )
+    ) +
+    scale_x_continuous(breaks = seq(0, 1, .5)) +
+    ggplot2::ylab(
+      label = databaseLabels[i]
+    ) +
+    ggplot2::xlab(
+      label = "Preference score"
+    ) +
+    ggplot2::scale_fill_manual(
+      values = alpha(c("#fc8d59", "#91bfdb"), .6)
+    ) +
+    ggplot2::scale_color_manual(
+      values = alpha(c("#fc8d59", "#91bfdb"), .9)
+    ) +
+    theme_bw() +
+    ggplot2::theme(
+      legend.title    = ggplot2::element_blank(),
+      legend.position = "none",
+      axis.title.x    = element_blank(),
+      axis.line.y     = element_blank(),
+      axis.ticks.y    = element_blank(),
+      axis.text.y     = element_blank(),
+      axis.text.x     = element_text(size = 22),
+      axis.title      = element_text(size = 30),
+      strip.text      = element_text(size = 25)
+    )  
+  if (i != 1) {
+    pp[[i]] <- pp[[i]] +
+      theme(
+        strip.text       = element_blank(),
+        strip.background = element_blank()
+      )
+  }
+}
+
+
+plot <- gridExtra::grid.arrange(pp[[1]], pp[[2]], pp[[3]], nrow = 1)
+ggsave(
+  "figures/OverallPsDensity.tiff",
+  plot, 
[…]

### 2021-12-03 code/plotMetaOverall.R

@@ -26,19 +26,7 @@ calibrateOverallResults <- readRDS(
     )
 )
 
-metaCalibrateOverall <- readRDS(
-    file.path(
-        "data/processed",
-        paste0(
-            paste(
-                "metaCalibrateOverall",
-                suffix,
-                sep = "_"
-            ),
-            ".rds"
-        )
-    )
-)
+metaCalibrateOverall <- readRDS("data/processed/metaCalibrateOverall.rds")
 
 overall <- calibrateOverallResults %>%
     select(-contains("Rr")) %>%
@@ -140,14 +128,16 @@ p <- ggplot(
         values = c(23, 21)
     ) +
     xlab("Calibrated hazard ratio") +
-    theme_classic() +
+    # theme_classic() +
+    ggthemes::theme_clean() +
     theme(
-        legend.position  = "none",
-        axis.title.y     = element_blank(),
-        strip.background = element_blank()
+      legend.position  = "none",
+      axis.title.y     = element_blank(),
+      strip.background = element_blank(),
+      strip.text.x     = element_text(size = 10)
     )
 
-ggsave("figures/plotMeta.pdf", plot = p, height = 3, width = 7)
-ggsave("figures/plotMeta.tiff", plot = p, height = 3, width = 7, compression = "lzw+p", dpi = 1000)
-ggsave("figures/plotMeta.png", plot = p, height = 3, width = 7)
+# ggsave("figures/plotMeta.pdf", plot = p, height = 3, width = 7)
+ggsave("figures/plotMeta.tiff", plot = p, height = 3, width = 7, compression = "lzw+p", dpi = 600)
+# ggsave("figures/plotMeta.png", plot = p, height = 3, width = 7)
 

### 2021-12-03 submission/manuscript.rmd
fix figure labels and themes; switch to cowplot

--- 2021-12-01 submission/manuscript.rmd
@@
 \newpage
-```{r plotMeta, echo=FALSE, fig.cap="Calibrated hazard ratios for the 3 outcomes of interest across the 3 considered databases using a set of 126 negative controls.
+```{r overallCovariateBalance, cache = TRUE, echo = FALSE, warning=FALSE, fig.cap="Calibrated overall results"}
+grid::grid.raster(tiff::readTIFF(here("figures/OverallCovariateBalance.tiff")))
+```
+
+\newpage
+```{r overallPsDensity, cache = FALSE, echo = FALSE, warning=FALSE, fig.cap="Calibrated overall results"}
+grid::grid.raster(tiff::readTIFF(here("figures/OverallPsDensity.tiff")))
+```
+
+\newpage
+```{r overallNcPlot, cache = FALSE, echo = FALSE, warning=FALSE, fig.cap="Overall negative control plot."}
+grid::grid.raster(tiff::readTIFF(here("figures/OverallNcPlot.tiff")))
+```
+
+\newpage
+```{r plotMeta, echo=FALSE, warning=FALSE, fig.cap="Calibrated hazard ratios for the 3 outcomes of interest across the 3 considered databases using a set of 126 negative controls.
 Values below 1 favour teriparatide, while values above 1 favour bisphosphonates.", out.width = '100%'}

### 2021-12-03 code/PlotCovariateBalance.R

@@ -17,30 +17,36 @@ for (i in seq_along(databases)) {
     bind_rows() %>%
     tibble() %>%	
     CohortMethod::plotCovariateBalanceScatterPlot(
-      beforeLabel = "",
-      afterLabel = databaseLabels[i]
-    ) +
+      beforeLabel = " ",
+      afterLabel = " "
+      ) +
+    ggtitle(databaseLabels[i]) +
     theme_bw() +
     theme(
-      plot.title   = element_blank(),
+      plot.margin  = margin(2, .25, 2, .25, "cm"),
+      plot.title   = element_text(size = 34),
       axis.title.x = element_blank(),
       axis.text.x  = element_text(size = 22),
       axis.text.y  = element_text(size = 22),
-      axis.title   = element_text(size = 30),
-      strip.text   = element_text(size = 25),
-      strip.background = element_rect(colour = "black", fill = NA)
+      axis.title   = element_text(size = 30)
     )
   
   if (i != 1) {
     pp[[i]] <- pp[[i]] +
       theme(
-        strip.text       = element_blank(),
-        strip.background = element_blank()
+        axis.text.y = element_blank()
       )
   }
 }
 
-plot <- gridExtra::grid.arrange(pp[[1]], pp[[2]], pp[[3]], nrow = 1)
+plot <- cowplot::plot_grid(
+                  pp[[1]],
+                  pp[[2]],
+                  pp[[3]],
+                  nrow = 1
+                ) +
+  cowplot::draw_label("Before matching", x = .5, y = 0, vjust = -.5, size = 30) +
+  cowplot::draw_label("After matching", x = 0, y = .5, vjust = 1.2, angle = 90, size = 30)
 ggsave(
   "figures/OverallCovariateBalance.tiff",
   plot, 

### 2021-12-03 code/PlotNegativeControls.R

@@ -41,14 +41,13 @@ for (i in seq_along(databases)) {
     logRrNegatives = negativeControls$logRr,
     seLogRrNegatives = negativeControls$seLogRr,
     logRrPositives = positiveResults$logRr,
-    seLogRrPositives = positiveResults$seLogRr
-  ) +
-    ggtitle(databaseLabels[i]) +
-    theme(
+    seLogRrPositives = positiveResults$seLogRr,
+    xLabel = ""
+  ) + ggtitle(databaseLabels[i]) + theme(
       plot.title = element_text(size = 38),
       axis.text.x = element_text(size = 22),
       axis.title.x  = element_text(size = 30),
-    )
+      )
   if (i == 1) {
     plot[[i]] <- plot[[i]] +
       theme(axis.title.y = element_text(size = 30))
@@ -57,8 +56,14 @@ for (i in seq_along(databases)) {
   }
 }
 
-res <- gridExtra::grid.arrange(plot[[1]], plot[[2]], plot[[3]], nrow = 1)
-
+# res <- gridExtra::grid.arrange(plot[[1]], plot[[2]], plot[[3]], nrow = 1)
+res <- cowplot::plot_grid(
+                  plot[[1]],
+                  plot[[2]],
+                  plot[[3]],
+                  nrow = 1
+                ) +
+  cowplot::draw_label("Relative risk", x = .5, y = 0, vjust = -.5, size = 30)
 
 ggsave(
   file.path("figures/OverallNcPlot.tiff"),

### 2021-12-03 code/PlotPsDensity.R

@@ -1,5 +1,10 @@
 #!/usr/bin/env Rscript
 
+# ===================================
+# File: PlotPsDensity.R
+# Description:
+# ===================================
+
 library(tidyverse)
 
 databases <- c("ccae", "optum_extended_dod", "optum_ehr")
@@ -34,23 +39,21 @@ for (i in seq_along(databases)) {
       )
     ) +
     scale_x_continuous(breaks = seq(0, 1, .5)) +
-    ggplot2::ylab(
-      label = databaseLabels[i]
-    ) +
-    ggplot2::xlab(
-      label = "Preference score"
-    ) +
     ggplot2::scale_fill_manual(
       values = alpha(c("#fc8d59", "#91bfdb"), .6)
     ) +
     ggplot2::scale_color_manual(
       values = alpha(c("#fc8d59", "#91bfdb"), .9)
     ) +
-    theme_bw() +
+    ggtitle(databaseLabels[i]) +
+    theme_classic() +
     ggplot2::theme(
+      plot.margin     = margin(2, .25, 2, .25, "cm"),
+      plot.title      = element_text(size = 34),
       legend.title    = ggplot2::element_blank(),
       legend.position = "none",
       axis.title.x    = element_blank(),
+      axis.title.y    = element_blank(),
       axis.line.y     = element_blank(),
       axis.ticks.y    = element_blank(),
       axis.text.y     = element_blank(),
@@ -58,17 +61,17 @@ for (i in seq_along(databases)) {
       axis.title      = element_text(size = 30),
       strip.text      = element_text(size = 25)
     )  
-  if (i != 1) {
-    pp[[i]] <- pp[[i]] +
-      theme(
-        strip.text       = element_blank(),
-        strip.background = element_blank()
-      )
-  }
 }
 
+plot <- cowplot::plot_grid(
+                  pp[[1]],
+                  pp[[2]],
+                  pp[[3]],
+                  nrow = 1
+                ) +
+  cowplot::draw_label("Preference score", x = .5, y = 0, vjust = -.5, size = 30)
+  
 
-plot <- gridExtra::grid.arrange(pp[[1]], pp[[2]], pp[[3]], nrow = 1)
 ggsave(
   "figures/OverallPsDensity.tiff",
   plot, 

### 2021-12-06 submission/manuscript.rmd
add risk stratified results and update methods

@@
 regression using the same predefined set of measured covariates
-[@Tian2018]. We estimated treatment effects using Cox proportional
-hazards models with treatment as the sole covariate fitted in the 1:4
-propensity score-matched subset of the considered patient
-population. Hazard ratios (HR) derived in different databases were
-summarized using random effects meta-analysis.
+[@Tian2018]. To assume that treatment and comaparator cohorts stand in
+equipoise we required that the majority of the patients in both have
+preference scores between 0.3 and 0.7. We assumed that covariate
+balance had been reached, if all after-adjustment baseline
+characteristics had standardized absolute mean differences of less
+than 0.1.  We estimated treatment effects using Cox proportional
+hazards models with treatment as the sole covariate. To account for
+observed confounding we used variable ratio matching (1:10) on the
+propensity score using a caliper of 0.2 on the logit scale.  Hazard
+ratios (HR) derived in different databases were summarized using
+random effects meta-analysis.
 
@@
 three efficacy outcomes. We built the prediction models using LASSO
-logistic regression on the propensity score matched (1:4)
+logistic regression on the propensity score matched (1:10)
 subpopulation of the pooled treatment arms, aiming to remove any
@@
 estimates were calculated based on the difference of the Kaplan-Meier
-estimates, on day 730 after treatment initiation.
+estimates, on day 365 after treatment initiation.
 
-For each outcome we used the derived prediction models to divide the
-population in 4 equally-sized subgroups, based on the quarters of the
-predicted risk distribution. Within each of these risk-based subgroups
-we developed a new propensity score model. Our analyses were performed
-on the propensity score matched (1:4) subset of the risk subgroup
-subset. We derived relative effect estimates using Cox proportional
-hazards models only with treatment as a predictor. Absolute effect
-estimates were calculated based on the difference of the Kaplan-Meier
-estimates, on day 730 after treatment initiation.
+For each outcome we used the derived prediction models to stratify the
+study population using two separate stratifcation schemes. In the
+first approach, we divide the population into the lowest 75\% and the
+top 25\% hip fracture risk subgroups. In the second approach, the
+population is divided using existing guidelines on teriparatide
+treatment based on major fracture risk (REF!!). Within each of these
+risk-based subgroups we developed a new propensity score model. Our
+analyses were performed on the propensity score matched (1:10) subset
+of the risk subgroup subset. We derived relative effect estimates
+using Cox proportional hazards models only with treatment as a
+predictor. Absolute effect estimates were calculated based on the
+difference of the Kaplan-Meier estimates, on day 365 after treatment
+initiation.
 
@@
 
+```{r, echo=FALSE, warning=FALSE, message=FALSE}
+metaCalibrateOverall <- readRDS(here("data/processed/metaCalibrateOverall.rds"))
+
+suffix <- "age_50_tr_1_q_25_75"
+
+calibrateOverallResults <- readRDS(
+    here(
+      file.path(
+          "data/processed",
+          paste0(
+              paste(
+                  "calibrateOverallResults",
+                  suffix,
+                  sep = "_"
+              ),
+              ".rds"
+          )
+      )
+    )
+)
+
+```
+
+In all databases adequate equipoise of the preference score
+distributions was achieved (Figure \ref{fig:overallPsDensity}). After
+matching all baseline covariates (>35,000 in each database) were well
+balanced (Figure \ref{fig:overallCovariateBalance}). This indicates
+that we were able account for the majority of the observed
+condfounding present in all databases. However, negative control
+analyses time falsely favored oral bisphosphonates over teriparatide,
+which suggests the presence of residual unobserved confoundng (Figure
+\ref{fig:overallNcPlot}).
+
+In terms of hip fracture risk we estimated calibrated hazard ratios of
+`r calibrateOverallResults %>% filter(database == "ccae", outcome == 101) %>% pull(hr)`
+(`r calibrateOverallResults %>% filter(database == "ccae", outcome == 101) %>% pull(lower)`
+to
+`r calibrateOverallResults %>% filter(database == "ccae", outcome == 101) %>% pull(upper)`;
+95\% CI), 
+`r calibrateOverallResults %>% filter(database == "optum_extended_dod", outcome == 101) %>% pull(hr)`
+(`r calibrateOverallResults %>% filter(database == "optum_extended_dod", outcome == 101) %>% pull(lower)`
+to
+`r calibrateOverallResults %>% filter(database == "optum_extended_dod", outcome == 101) %>% pull(upper)`;
+95\% CI) and
+`r calibrateOverallResults %>% filter(database == "optum_ehr", outcome == 101) %>% pull(hr)`
+(`r calibrateOverallResults %>% filter(database == "optum_ehr", outcome == 101) %>% pull(lower)`
+to
+`r calibrateOverallResults %>% filter(database == "optum_ehr", outcome == 101) %>% pull(upper)`;
+95\% CI), in CCAE, Optum-DOD and Optum-EHR respectively. The
+calibrated meta-analytic hazard ratio of the overall effect of
+teriparatide compared to oral bisphosphonates was
+`r metaCalibrateOverall %>% filter(outcome == 101) %>% pull(hr)`
+(`r metaCalibrateOverall %>% filter(outcome == 101) %>% pull(lower)`
+to
+`r metaCalibrateOverall %>% filter(outcome == 101) %>% pull(upper)`;
+95\% CI). Results for major osteoporotic and vertebral fractures are
+provided in the supplement.
+
+When using the internally developed hip fracture risk prediction models to divide the population into lower 75\% and top 25\% risk subgroups, we observed....
+
 \newpage
@@
 # Tables and figures
+\newpage
+```{r overallPsDensity, cache = TRUE, echo = FALSE, warning=FALSE, fig.cap="Calibrated overall results"}
+grid::grid.raster(tiff::readTIFF(here("figures/OverallPsDensity.tiff")))
+```
 
@@
 \newpage
-```{r overallPsDensity, cache = FALSE, echo = FALSE, warning=FALSE, fig.cap="Calibrated overall results"}
-grid::grid.raster(tiff::readTIFF(here("figures/OverallPsDensity.tiff")))
-```
-
-\newpage
-```{r overallNcPlot, cache = FALSE, echo = FALSE, warning=FALSE, fig.cap="Overall negative control plot."}
+```{r overallNcPlot, cache = TRUE, echo = FALSE, warning=FALSE, fig.cap="Overall negative control plot."}
 grid::grid.raster(tiff::readTIFF(here("figures/OverallNcPlot.tiff")))
@@
 \newpage
-```{r plotMeta, echo=FALSE, warning=FALSE, fig.cap="Calibrated hazard ratios for the 3 outcomes of interest across the 3 considered databases using a set of 126 negative controls.
+```{r plotMeta, cache=TRUE, echo=FALSE, warning=FALSE, fig.cap="Calibrated hazard ratios for the 3 outcomes of interest across the 3 considered databases using a set of 126 negative controls.
 Values below 1 favour teriparatide, while values above 1 favour bisphosphonates.", out.width = '100%'}
@@
 ```
+
+\newpage
+```{r plotMetaRiskStratified, cache=TRUE, echo=FALSE, warning=FALSE, fig.cap="Calibrated hazard ratios for the 3 outcomes of interest across the 3 considered databases using a set of 126 negative controls.
+Values below 1 favour teriparatide, while values above 1 favour bisphosphonates.", out.width = '100%'}
+grid::grid.raster(tiff::readTIFF(here::here("figures/plotMetaRiskStratified.tiff")))
+```
+
+\newpage
+```{r plotAbsoluteRiskStratified, cache=FALSE, echo=FALSE, warning=FALSE, fig.cap="Calibrated hazard ratios for the 3 outcomes of interest across the 3 considered databases using a set of 126 negative controls.
+Values below 1 favour teriparatide, while values above 1 favour bisphosphonates.", out.width = '100%'}
+grid::grid.raster(tiff::readTIFF(here::here("figures/plotAbsoluteHip.tiff")))
+```

### 2021-12-06 code/plotAbsoluteHip.R

@@ -6,7 +6,7 @@
 #   Plots the risk stratified absolute difference for hip fracture with respect
 #   to estimated hip fracture risk
 # Depends:
-#   data/raw/mappedOverallAbsoluteResults.rds
+#   data/processed/mappedOverallAbsoluteResults.rds
 # Output:
 #   figures/plotAbsoluteHip.pdf
 #   figures/plotAbsoluteHip.tiff
@@ -17,26 +17,25 @@
 
 library(tidyverse)
 
-absolute <- readRDS("data/raw/mappedOverallAbsoluteResults.rds") %>%
+absolute <- readRDS("data/processed/mappedOverallAbsoluteResults.rds") %>%
     filter(
-        analysisType == "matchOnPs_1_to_4",
+        analysisType == "age_50_tr_1_q_25_75",
         stratOutcome == 101,
-        estOutcome   == 101
+        estOutcome   == 101,
+        database != "mdcr"
     ) %>%
     mutate(
         estimate = 100 * estimate,
         lower    = 100 * lower,
         upper    = 100 * upper,
         meanRisk = case_when(
-            riskStratum == "Q1" ~1,
-            riskStratum == "Q2" ~2,
-            riskStratum == "Q3" ~3,
-            riskStratum == "Q4" ~4
+            riskStratum == "Q1" ~.5,
+            riskStratum == "Q2" ~1.5
         ),
         database = case_when(
-            database == "mdcr" ~ "MDCR",
-            database == "optum_dod" ~ "Optum-DOD",
-            database == "panther" ~ "Optum-EHR"
+            database == "ccae" ~ "CCAE",
+            database == "optum_extended_dod" ~ "Optum-DOD",
+            database == "optum_ehr" ~ "Optum-EHR"
         )
     )
 
@@ -51,20 +50,8 @@ dataRect <- function(xmin, xmax) {
 
 p <- ggplot() +
     geom_rect(
-        data = dataRect(-Inf, 1.5),
+        data = dataRect(-Inf, 1),
         inherit.aes = FALSE,
-        aes(
-            xmin = xmin,
-            xmax = xmax,
-            ymin = ymin,
-            ymax = ymax
-        ),
-        fill = "#ffffcc",
-        alpha = .3
-    ) +
-    geom_rect(
-        data = dataRect(1.5, 2.5),
-        inherit.aes = F,
         aes(
             xmin = xmin,
             xmax = xmax,
@@ -75,19 +62,7 @@ p <- ggplot() +
[…]

### 2021-12-06 code/plotMetaRiskStratified.R

@@ -26,7 +26,8 @@ calibrateRiskStratified <- readRDS(
     )
 ) %>%
     filter(
-        stratOutcome == 101
+      stratOutcome == 101,
+      estOutcome == 101
     ) %>%
     rename("outcome" = "estOutcome")
 
@@ -43,7 +44,7 @@ metaCalibrateRiskStratified <- readRDS(
         )
     )
 ) %>%
-    filter(stratOutcome == 101) %>%
+    filter(stratOutcome == 101, estOutcome == 101) %>%
     rename("outcome" = "estOutcome")
 
 riskStratified <- calibrateRiskStratified %>%
@@ -72,20 +73,19 @@ metaRiskStratified <- metaCalibrateRiskStratified %>%
 
 combined <- rbind(riskStratified, metaRiskStratified) %>%
     mutate(
-        database = factor(
-            x = database,
-            levels = c("overall", "optum_extended_dod", "optum_ehr", "ccae"),
-            labels = c("Overall", "OPTUM (DoD)", "Panther", "CCAE")
-        ),
-        outcome = factor(
-            x = outcome,
-            levels = 101:103,
-            labels = c(
-                "Hip fracture",
-                "Major osteoporotic fracture",
-                "Vertebral fracture"
-            )
+      database = factor(
+        x = database,
+        levels = c("overall", "optum_extended_dod", "optum_ehr", "ccae"),
+        labels = c("Overall", "OPTUM (DoD)", "Panther", "CCAE")
+      ),
+      riskStratum = factor(
+        x = riskStratum,
+        levels = c("Q1", "Q2"),
+        labels = c(
+          "Lower 75%\nhip fracture risk",
+          "Upper 25%\nhip fracture risk"
         )
+      )
     )
 
 p <- ggplot(
@@ -99,7 +99,7 @@ p <- ggplot(
     )
 ) +
     facet_wrap(
-        ~riskStratum + outcome
+        ~riskStratum 
     ) +
     geom_point(
         aes(
@@ -119,7 +119,7 @@ p <- ggplot(
     ) +
     geom_text(
         label = "Favors Bisphosphonates", 
-        x     = 1.675, 
+        x     = 1.555, 
         y     = -.1, 
         color = "black",
         size  = 2
@@ -137,7 +137,7 @@ p <- ggplot(
     scale_x_continuous(
         breaks = c(0, 1, 2),
         labels = c("0", "1", "2"),
-        limits = c(0, 3)
+        limits = c(0, 2.2)
     ) +
[…]

### 2021-12-22 submission/manuscript.rmd
start table 1

@@
 
+```{r, echo=FALSE, warning=FALSE, message=FALSE}
+
+ccae <- tibble(
+  database = "ccae",
+  variable = c(
+    "Osteoarthritis",
+    "Rheumatoid arthritis",
+    "Renal impairment",
+    "Crohn's disease",
+    "Obesity",
+    "Antithrombotic agents",
+    "Drugs used in diabetes",
+    "Lipid modifying agents",
+    "Drugs for obstructive airway diseases",
+    "Immunosuppressants",
+    "Antineoplastic agents",
+    "Antiinflammatory and antireumatic agents",
+    "Opioids",
+    "Antiepileptics",
+    "Psycholeptics",
+    "Antidepressants"
+  ),
+  treatment = c(
+    41.7, 7.6, 4, 1.4, 4.8, 11.7, 7.2, 26.8, 44.3, 12.1,
+    11.8, 36.8, 43.3, 23.7, 40.7, 39.2
+  ),
+  comparator = c(
+    23.3, 4.1, 1.7, .7,
+    2.9, 6, 6.3, 29.2,
+    34.2, 6.2, 8.6, 28.4,
+    26.8, 12, 28.3, 28.4
+  ),
+  std_difference = c(
+    -.23, -.1, -.09, -.25,
+    -.07, -.14, -.03, .03,
+    -.11, -.14, -.07, -.1,
+    -.2, -.2, -.15, -.13
+  )
+)
+
+optum_ehr <- tibble(
+  database = "optum-ehr",
+  variable = c(
+    "Osteoarthritis",
+    "Rheumatoid arthritis",
+    "Renal impairment",
+    "Crohn's disease",
+    "Obesity",
+    "Antithrombotic agents",
+    "Drugs used in diabetes",
+    "Lipid modifying agents",
+    "Drugs for obstructive airway diseases",
+    "Immunosuppressants",
+    "Antineoplastic agents",
+    "Antiinflammatory and antireumatic agents",
+    "Opioids",
+    "Antiepileptics",
+    "Psycholeptics",
+    "Antidepressants"
+  ),
+  treatment = c(
+    39.6, 7.7, 8.3,
+    1, 5.9, 41.4, 11.3, 35.6,
+    48.4, 11.7, 12, 55.7, 49.7,
+    28, 44.3, 39.2 
+  ),
+  comparator = c(
+    30.3, 4.7, 8.4, .6, 6.2,
+    38.5, 12.4, 42.3, 42.2, 6.6,
+    8.2, 50.5, 33.4, 17.3, 34.5, 31
+  ),
+  std_difference = c(
+    -.11, -.09, 0, -.03, .01,
+    -.03, .02, .08, -.07, -.12,
+    -.09, -.05, -.18, -.16, -.11, -.1
+  )
+)
+
+optum_dod <- tibble(
+  database = "optum-dod",
+  variable = c(
+    "Osteoarthritis",
+    "Rheumatoid arthritis",
+    "Renal impairment",
+    "Crohn's disease",
+    "Obesity",
+    "Antithrombotic agents",
+    "Drugs used in diabetes",
+    "Lipid modifying agents",
+    "Drugs for obstructive airway diseases",
+    "Immunosuppressants",
+    "Antineoplastic agents",
+    "Antiinflammatory and antireumatic agents",
+    "Opioids",
+    "Antiepileptics",
+    "Psycholeptics",
+    "Antidepressants"
+  ),
+  treatment = c(
+    56.6, 10.3, 12.7, 1.1, 7.3,
+    18.9, 11.1, 38.5, 48.1, 12.8,
+    13.9, 37.3, 50.1, 27.2, 40.2, 41.1
+  ),
+  comparator = c(
+    38.6, 5.3, 9.4, .5, 5.9,
+    12.6, 11.8, 43.1, 37, 5.8,
+    8.9, 28.2, 29.6, 15, 26.8, 29
+  ),
+  std_difference = c(
+    -.18, -.13, -.07, -.04, -.04,
+    -.11, .02, .05, -.12, -.16,
+    -.11, -.11, -.23, -.15, -.16, -.15
+  )
+)
+
+table1 <- bind_rows(ccae, optum_dod, optum_ehr)
+
+table1 %>%
+  pivot_wider(
+    names_from = database, 
+    values_from = c(treatment, comparator, std_difference)
+  ) %>% 
+  relocate(
+    "variable", 
+    "treatment_ccae", 
+    "comparator_ccae", 
+    "std_difference_ccae",
+    "treatment_optum-dod",
+    "comparator_optum-dod", 
+    "std_difference_optum-dod"
+) %>%
+kbl(
+  format = "latex",
+  align = "r",
+  booktabs = TRUE,
+  col.names = c(
+    "Variable",
+    rep(
+      c("Teriparatide", "Bisphosphonates", "Std. diff."),
+      3
+    )
+  )
+) %>%
+add_header_above(
+  c(" " = 1, "CCAE" = 3, "OPTUM-DOD" = 3, "OPTUM-EHR" = 3)
+)
+```
+
 ***[ Introductory stuff for Table 1 !! ]***
@@
 
-When using the internally developed hip fracture risk prediction models to divide the population into lower 75\% and top 25\% risk subgroups, we observed....
+```{r, echo=FALSE, warning=FALSE, message=FALSE}
+suffix <- "age_50_tr_1_q_25_75"
+
+calibrateRiskStratified <- readRDS(
+    file.path(
+        here(
+          "data/processed",
+          paste0(
+              paste(
+                  "calibrateRiskStratified",
+                  suffix,
+                  sep = "_"
+              ),
+              ".rds"
+          )
+        )
+    )
+) %>%
+    filter(
+      stratOutcome == 101,
+      estOutcome == 101
+    ) %>%
+    rename("outcome" = "estOutcome")
+
+metaCalibrateRiskStratified <- readRDS(
+    file.path(
+      here(
+        "data/processed",
+        paste0(
+            paste(
+                "metaCalibrateRiskStratified",
+                suffix,
+                sep = "_"
+            ),
+            ".rds"
+        )
+    )
+  )
+) %>%
+    filter(stratOutcome == 101, estOutcome == 101) %>%
+    rename("outcome" = "estOutcome")
+
+riskStratified <- calibrateRiskStratified %>%
+    mutate(
+        hr = exp(logRr),
+        lower    = exp(logLb95Rr),
+        upper    = exp(logUb95Rr)
+    ) %>%
+    select(-contains("Rr")) %>%
+    mutate(
+        type = "single",
+        position = case_when(
+            database == "ccae" ~ 5,
+            database == "optum_extended_dod" ~ 4,
+            database == "optum_ehr" ~ 3
+        )
+    )
+
+metaRiskStratified <- metaCalibrateRiskStratified %>%
+    mutate(
+        database = "overall",
+        type     = "meta",
+        position = 1
+    ) %>%
+    relocate(database, .after = stratOutcome)
+
+combined <- rbind(riskStratified, metaRiskStratified)
+
+absolute <- readRDS(here("data/processed/mappedOverallAbsoluteResults.rds")) %>%
+    filter(
+        analysisType == "age_50_tr_1_q_25_75",
+        stratOutcome == 101,
+        estOutcome   == 101,
+        database != "mdcr"
+    ) %>%
+    mutate(
+        estimate = 100 * estimate,
+        lower    = 100 * lower,
+        upper    = 100 * upper
+    )
+```
+
+When using the internally developed hip fracture risk prediction
+models to divide the population into lower 75\% and top 25\% risk
+subgroups, we estimated calibrated hazard ratios of 
+`r combined %>% filter(database == "ccae", riskStratum == "Q1") %>% pull(hr)`
+(`r combined %>% filter(database == "ccae", riskStratum == "Q1") %>% pull(lower)`
+to
+`r combined %>% filter(database == "ccae", riskStratum == "Q1") %>% pull(upper)`;
+95\% CI),
+`r combined %>% filter(database == "optum_extended_dod", riskStratum == "Q1") %>% pull(hr)`
+(`r combined %>% filter(database == "optum_extended_dod", riskStratum == "Q1") %>% pull(lower)`
+to
+`r combined %>% filter(database == "optum_extended_dod", riskStratum == "Q1") %>% pull(upper)`;
+95\% CI) and
+`r combined %>% filter(database == "optum_ehr", riskStratum == "Q1") %>% pull(hr)`
+(`r combined %>% filter(database == "optum_ehr", riskStratum == "Q1") %>% pull(lower)`
+to
+`r combined %>% filter(database == "optum_ehr", riskStratum == "Q1") %>% pull(upper)`;
+95\% CI) in the lower 75\% hip fracture risk subgroup in CCAE,
+Optum-DOD and Optum-EHR, respectively. These translated to absolute risk reductions of 
+`r absolute %>% filter(riskStratum == "Q1", database == "ccae") %>% pull(estimate)`
+(`r absolute %>% filter(riskStratum == "Q1", database == "ccae") %>% pull(lower)`
+to
+`r absolute %>% filter(riskStratum == "Q1", database == "ccae") %>% pull(upper)`;
+95\% CI), 
+`r absolute %>% filter(riskStratum == "Q1", database == "optum_extended_dod") %>% pull(estimate)`
+(`r absolute %>% filter(riskStratum == "Q1", database == "optum_extended_dod") %>% pull(lower)`
+to
+`r absolute %>% filter(riskStratum == "Q1", database == "optum_extended_dod") %>% pull(upper)`;
+95\% CI) and
+`r absolute %>% filter(riskStratum == "Q1", database == "optum_ehr") %>% pull(estimate)`
+(`r absolute %>% filter(riskStratum == "Q1", database == "optum_ehr") %>% pull(lower)`
+to
+`r absolute %>% filter(riskStratum == "Q1", database == "optum_ehr") %>% pull(upper)`;
+95\% CI) in CCAE, Optum-DOD and Optum-EHR respectively. Regarding hip
+fracture risk, the overall meta-analytic hazard ratio of the effect of
+teriparatide compared to oral bisphosphonates in the lower 75\% hip
+fracture risk subgroup was
+`r combined %>% filter(database == "overall", riskStratum == "Q1") %>% pull(hr)`
+(`r combined %>% filter(database == "overall", riskStratum == "Q1") %>% pull(lower)`
+to
+`r combined %>% filter(database == "overall", riskStratum == "Q1") %>% pull(upper)`;
+95\% CI) across all 3 databases in the lower risk subgroup.
+
+In the upper 25\% hip fracture risk subgroup we estimated hazard ratios of
+`r combined %>% filter(database == "ccae", riskStratum == "Q2") %>% pull(hr)`
+(`r combined %>% filter(database == "ccae", riskStratum == "Q2") %>% pull(lower)`
+to
+`r combined %>% filter(database == "ccae", riskStratum == "Q2") %>% pull(upper)`;
+95\% CI),
+`r combined %>% filter(database == "optum_extended_dod", riskStratum == "Q2") %>% pull(hr)`
+(`r combined %>% filter(database == "optum_extended_dod", riskStratum == "Q2") %>% pull(lower)`
+to
+`r combined %>% filter(database == "optum_extended_dod", riskStratum == "Q2") %>% pull(upper)`;
+95\% CI) and
+`r combined %>% filter(database == "optum_ehr", riskStratum == "Q2") %>% pull(hr)`
+(`r combined %>% filter(database == "optum_ehr", riskStratum == "Q2") %>% pull(lower)`
+to
+`r combined %>% filter(database == "optum_ehr", riskStratum == "Q2") %>% pull(upper)`;
+95\% CI) in the lower 75\% hip fracture risk subgroup in CCAE,
+Optum-DOD and Optum-EHR, respectively. These translated to absolute risk reductions of 
+`r absolute %>% filter(riskStratum == "Q2", database == "ccae") %>% pull(estimate)`
+(`r absolute %>% filter(riskStratum == "Q2", database == "ccae") %>% pull(lower)`
+to
+`r absolute %>% filter(riskStratum == "Q2", database == "ccae") %>% pull(upper)`;
+95\% CI), 
+`r absolute %>% filter(riskStratum == "Q2", database == "optum_extended_dod") %>% pull(estimate)`
+(`r absolute %>% filter(riskStratum == "Q2", database == "optum_extended_dod") %>% pull(lower)`
+to
+`r absolute %>% filter(riskStratum == "Q2", database == "optum_extended_dod") %>% pull(upper)`;
+95\% CI) and
+`r absolute %>% filter(riskStratum == "Q2", database == "optum_ehr") %>% pull(estimate)`
+(`r absolute %>% filter(riskStratum == "Q2", database == "optum_ehr") %>% pull(lower)`
+to
+`r absolute %>% filter(riskStratum == "Q2", database == "optum_ehr") %>% pull(upper)`;
+95\% CI) in CCAE, Optum-DOD and Optum-EHR respectively. Regarding hip
+fracture risk, the overall meta-analytic hazard ratio of the effect of
+teriparatide compared to oral bisphosphonates in the lower 75\% hip
+fracture risk subgroup was
+`r combined %>% filter(database == "overall", riskStratum == "Q2") %>% pull(hr)`
+(`r combined %>% filter(database == "overall", riskStratum == "Q2") %>% pull(lower)`
+to
+`r combined %>% filter(database == "overall", riskStratum == "Q2") %>% pull(upper)`;
+95\% CI) across all 3 databases in the upper risk subgroup.
 
